---
title: "FoodSeq Reference Database Pipeline"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
---

This notebook builds three reference databases:

- **Part A — trnL g/h**: Plant sequences amplified by the trnLg/trnLh primer pair, combining RefSeq plastid genomes and GenBank
- **Part A-CD — trnL c/d**: Plant sequences amplified by the trnLc/trnLd primer pair (the full trnL intron), re-extracted from Part A's inputs without additional NCBI queries
- **Part B — 12SV5**: Vertebrate dietary sequences amplified by the 12SV5 primer pair, combining RefSeq mitochondrial genomes and GenBank

Both pipelines share the same setup, SLURM jobs, and taxonomy approach.

**Before running:** update the paths in the Configuration chunk below to match your cluster environment.

---

# Setup

## Configuration

Update these paths and parameters before running anything else. All other chunks reference these variables.

```{r configuration}
# ── Cluster paths ─────────────────────────────────────────────────────────────
SCRATCH      <- "/scratch/your_username"                  # cluster scratch directory
REPO_DIR     <- "/path/to/food-dbs"                       # cloned repo root
PLASTID_DIR  <- file.path(SCRATCH, "plastid_refseq")      # RefSeq plastid download
MITO_DIR     <- file.path(SCRATCH, "mito_refseq")         # RefSeq mitochondrial download
SQL_PATH     <- file.path(SCRATCH, "accessionTaxa.sql")   # taxonomizr SQL database

# ── SLURM parameters ──────────────────────────────────────────────────────────
PARTITION    <- "general"    # partition/queue name; check with 'sinfo'
R_MODULE     <- "R/4.3.0"    # R module name; check with 'module avail R'

# ── Output directories ────────────────────────────────────────────────────────
TRNL_DADA2    <- file.path(REPO_DIR, "data", "outputs", "dada2-compatible", "trnL")
TRNL_QIIME2   <- file.path(REPO_DIR, "data", "outputs", "qiime2-compatible", "trnL")
TRNLCD_DADA2  <- file.path(REPO_DIR, "data", "outputs", "dada2-compatible", "trnLCD")
TRNLCD_QIIME2 <- file.path(REPO_DIR, "data", "outputs", "qiime2-compatible", "trnLCD")
S12V5_DADA2   <- file.path(REPO_DIR, "data", "outputs", "dada2-compatible", "12Sv5")
S12V5_QIIME2  <- file.path(REPO_DIR, "data", "outputs", "qiime2-compatible", "12Sv5")

for (d in c(TRNL_DADA2, TRNL_QIIME2, TRNLCD_DADA2, TRNLCD_QIIME2,
            S12V5_DADA2, S12V5_QIIME2)) {
    dir.create(d, recursive = TRUE, showWarnings = FALSE)
}

cat("Repo:   ", REPO_DIR, "\n")
cat("Scratch:", SCRATCH,  "\n")
```

## Install packages

Run once on first use; safe to skip on subsequent runs.

```{r install, eval=FALSE}
# Bioconductor packages. pwalign holds pairwiseAlignment() on Bioconductor
# >= 3.19 (Part A-CD needs it); on older installations it is part of
# Biostrings and this install line simply skips it.
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(c("Biostrings", "ShortRead", "pwalign"), ask = FALSE)

# CRAN packages
install.packages(c("taxonomizr", "tidyverse", "rentrez", "remotes"),
                 repos = "https://cloud.r-project.org")

# GitHub packages
remotes::install_github("ammararuby/MButils")
```

---

# SLURM jobs

Submit all three jobs before proceeding to Part A or B. They run in parallel and
will each take several hours. Monitor progress in the following section.

## Download RefSeq plastid sequences (for trnL)

```{r submit_plastid}
plastid_script <- paste0(
"#!/bin/bash\n",
"#SBATCH --job-name=refseq_plastid\n",
"#SBATCH --partition=", PARTITION, "\n",
"#SBATCH --time=04:00:00\n",
"#SBATCH --mem=8G\n",
"#SBATCH --cpus-per-task=2\n",
"#SBATCH --output=", SCRATCH, "/refseq_plastid.log\n",
"\n",
"mkdir -p ", PLASTID_DIR, "\n",
"wget -r -nd -A '*.genomic.fna.gz' \\\n",
"  ftp://ftp.ncbi.nlm.nih.gov/refseq/release/plastid/ \\\n",
"  -P ", PLASTID_DIR, "\n",
"zcat ", PLASTID_DIR, "/*.genomic.fna.gz > ", PLASTID_DIR, "/refseq_plastid_all.fasta\n",
"echo 'Plastid download complete.'\n"
)

plastid_sh <- file.path(SCRATCH, "download_plastid.sh")
writeLines(plastid_script, plastid_sh)

cat("Script written to:", plastid_sh, "\n")
cat("Submit from the OOD terminal with:\n")
cat("  sbatch", plastid_sh, "\n")
```

## Download RefSeq mitochondrial sequences (for 12SV5)

```{r submit_mito}
mito_script <- paste0(
"#!/bin/bash\n",
"#SBATCH --job-name=refseq_mito\n",
"#SBATCH --partition=", PARTITION, "\n",
"#SBATCH --time=04:00:00\n",
"#SBATCH --mem=8G\n",
"#SBATCH --cpus-per-task=2\n",
"#SBATCH --output=", SCRATCH, "/refseq_mito.log\n",
"\n",
"mkdir -p ", MITO_DIR, "\n",
"wget -r -nd -A '*.genomic.fna.gz' \\\n",
"  ftp://ftp.ncbi.nlm.nih.gov/refseq/release/mitochondrion/ \\\n",
"  -P ", MITO_DIR, "\n",
"zcat ", MITO_DIR, "/*.genomic.fna.gz > ", MITO_DIR, "/refseq_mito_all.fasta\n",
"echo 'Mitochondrial download complete.'\n"
)

mito_sh <- file.path(SCRATCH, "download_mito.sh")
writeLines(mito_script, mito_sh)

cat("Script written to:", mito_sh, "\n")
cat("Submit from the OOD terminal with:\n")
cat("  sbatch", mito_sh, "\n")
```

## Build taxonomizr SQL database (shared by both pipelines)

```{r submit_taxonomizr}
taxonomizr_script <- paste0(
"#!/bin/bash\n",
"#SBATCH --job-name=taxonomizr_build\n",
"#SBATCH --partition=", PARTITION, "\n",
"#SBATCH --time=12:00:00\n",
"#SBATCH --mem=16G\n",
"#SBATCH --cpus-per-task=4\n",
"#SBATCH --output=", SCRATCH, "/taxonomizr_build.log\n",
"\n",
"module load ", R_MODULE, "\n",
"\n",
"Rscript -e \"\n",
"  library(taxonomizr)\n",
"  prepareDatabase('", SQL_PATH, "')\n",
"  cat('taxonomizr database built.\\n')\n",
"\""
)

tax_sh <- file.path(SCRATCH, "build_taxonomizr.sh")
writeLines(taxonomizr_script, tax_sh)

cat("Script written to:", tax_sh, "\n")
cat("Submit from the OOD terminal with:\n")
cat("  sbatch", tax_sh, "\n")
```

## Monitor jobs

Re-run this chunk to check status. Proceed to Part A or B once the relevant jobs
are no longer shown in the queue.

```{r monitor}
# Check job status from the OOD terminal with: squeue --me
# Log files are checked here automatically as jobs run

logs <- list(
    plastid    = file.path(SCRATCH, "refseq_plastid.log"),
    mito       = file.path(SCRATCH, "refseq_mito.log"),
    taxonomizr = file.path(SCRATCH, "taxonomizr_build.log")
)

for (name in names(logs)) {
    if (file.exists(logs[[name]])) {
        lines <- readLines(logs[[name]])
        cat("\n---", name, "---\n")
        cat("  Lines logged:", length(lines), "\n")
        cat("  Last 3 lines:\n")
        cat(paste(" ", tail(lines, 3), collapse = "\n"), "\n")
        completed <- any(grepl("complete|done|finished", lines, ignore.case = TRUE))
        cat("  Status:", ifelse(completed, "COMPLETE", "IN PROGRESS"), "\n")
    } else {
        cat("\n---", name, "---\n")
        cat("  Log not yet created (job not started or still queued)\n")
    }
}
```

---

# Shared libraries and functions

Run this before starting either Part A or Part B.

```{r libraries, message=FALSE, warning=FALSE}
library(Biostrings)
library(ShortRead)
library(taxonomizr)
library(tidyverse)
library(rentrez)
# Used for lowest_level() in A7 / CD7 / B7. Loaded here rather than only
# called as MButils::lowest_level() so a broken install fails now, at setup,
# instead of hours into the run. MButils imports dada2, so if this errors with
# an undefined symbol (e.g. _ZTIN3tbb4taskE), rebuild dada2 from source
# against the current RcppParallel:
#   BiocManager::install("dada2", type = "source", force = TRUE)
library(MButils)

source(file.path(REPO_DIR, "code", "functions", "find_primer_pair.R"))
source(file.path(REPO_DIR, "code", "functions", "query_ncbi.R"))
source(file.path(REPO_DIR, "code", "functions", "query_ncbi_accession.R"))
```

## NCBI API key

Setting an API key increases the NCBI e-utils rate limit from 3 to 10 requests per
second, which significantly speeds up the remote queries in sections A4 and B4.
Create an NCBI account, go to **Account Settings → API Key Management**, copy your
key, and paste it below. Also note that NCBI recommends running large queries on
weekends or between **9 PM and 5 AM EST** on weekdays.

```{r ncbi_api_key, eval=FALSE}
# Paste your NCBI API key here (keep the quotes)
rentrez::set_entrez_key("your_api_key_here")
```

```{r plotting, include=FALSE}
theme_set(
    theme_bw() +
    theme(
        axis.text  = element_text(size = 12),
        axis.title = element_text(size = 14, face = "bold"),
        strip.text = element_text(size = 12, face = "bold")
    )
)
```

```{r extract_taxonomy_fn}
# Shared helper: converts a vector of taxon IDs to a full taxonomy data frame
extract_taxonomy <- function(ids, sql_path) {
    # NCBI reclassified Eukaryota/Bacteria/Archaea from rank "superkingdom" to
    # rank "domain" (taxonomizr >= 0.11 follows suit). getRawTaxonomy() returns
    # the top-level rank under whichever name the underlying taxdump used, so we
    # capture BOTH names here and merge them into "superkingdom" below. Without
    # this, a database built from a current taxdump yields "domain" and the
    # requested "superkingdom" column comes back 100% NA.
    vars <- c("superkingdom", "domain", "phylum", "class", "order",
              "family", "genus", "species", "subspecies", "varietas", "forma")

    taxonomy.raw <- getRawTaxonomy(ids, sql_path)

    taxonomy <- as.data.frame(
        setNames(lapply(vars, function(v) character(0)), vars),
        stringsAsFactors = FALSE
    )

    for (i in seq_along(taxonomy.raw)) {
        row.i  <- taxonomy.raw[[i]] |> t() |> data.frame()
        shared <- intersect(vars, names(row.i))
        row.i  <- select(row.i, one_of(shared))
        taxonomy <- bind_rows(taxonomy, row.i)
    }

    # Merge "domain" into "superkingdom" so the top-level rank is populated
    # regardless of which name the taxdump used, then drop the helper column.
    taxonomy$superkingdom <- coalesce(taxonomy$superkingdom, taxonomy$domain)
    taxonomy$domain <- NULL

    taxonomy$taxid <- names(taxonomy.raw) |> trimws() |> as.integer()
    select(taxonomy, taxid, everything())
}
```

---

# Part A — trnL Reference Database

Primers: trnLg (`GGGCAATCCTGAGCCAA`) / trnLh (`CCATTGAGTCTCTGCACCTATC`)
Source organisms: curated food plant list (`human-foods.csv`)

## A1. Load inputs

```{r trnl_inputs}
trnLG <- DNAString("GGGCAATCCTGAGCCAA")
trnLH <- DNAString("CCATTGAGTCTCTGCACCTATC")
trnL_primers <- list(trnLG, trnLH)

plants <-
    file.path(REPO_DIR, "data", "inputs", "human-foods.csv") |>
    read.csv(stringsAsFactors = FALSE) |>
    filter(category == "plant") |>
    pull(scientific_name)

edits <-
    file.path(REPO_DIR, "data", "inputs", "Manual renaming.csv") |>
    read_csv()

cat(length(plants), "food plant species\n")
head(plants)
```

## A2. Filter RefSeq plastid to one sequence per species

```{r trnl_refseq_filter}
# "_acc" filename: names are "<accession> <Genus species>". Do NOT fall back to
# an old refseq_plastid_species.fasta cache — its names are species-only, which
# put a bare genus in the accession field of every RefSeq record (see B2 for the
# full failure chain this caused in the May 2026 12S build).
plastid_species_path <- file.path(PLASTID_DIR, "refseq_plastid_species_acc.fasta")

if (!file.exists(plastid_species_path)) {
    cat("Filtering RefSeq plastid to one sequence per species...\n")
    plastid <- readDNAStringSet(file.path(PLASTID_DIR, "refseq_plastid_all.fasta"))
    cat(length(plastid), "total sequences\n")

    accession       <- sub("^(\\S+).*", "\\1", names(plastid))
    species         <- sub("^\\S+\\s+(\\S+\\s+\\S+).*", "\\1", names(plastid))
    plastid_species <- plastid[!duplicated(species)]
    names(plastid_species) <- paste(accession, species)[!duplicated(species)]

    writeXStringSet(plastid_species, plastid_species_path)
    cat(length(plastid_species), "species retained\n")
} else {
    plastid_species <- readDNAStringSet(plastid_species_path)
    cat(length(plastid_species), "species loaded from existing file\n")
}

plastid_species
```

## A3. RefSeq: in silico PCR and subset to food plants

```{r trnl_refseq_pcr}
refseq.trnL <- find_primer_pair(plastid_species,
                                fwd = trnL_primers[[1]],
                                rev = trnL_primers[[2]])
cat(length(refseq.trnL), "sequences contain the trnL g/h primer pair\n")

plants.i    <- lapply(plants, grep, x = names(refseq.trnL)) |> unlist()
refseq.trnL <- refseq.trnL[plants.i]
cat(length(refseq.trnL), "food plant trnL sequences from RefSeq\n")

names(refseq.trnL) <- gsub(" .*$", "", names(refseq.trnL))
head(names(refseq.trnL))
```

## A4. NCBI GenBank: remote query

```{r trnl_ncbi_query}
cat("Querying NCBI for trnL sequences... this may take 10-30 minutes\n")
ncbi.trnL <- query_ncbi(marker = "trnL", organisms = plants)
cat(length(ncbi.trnL), "sequences returned from NCBI\n")
```

```{r trnl_ncbi_pcr}
# Keep the untrimmed pull for Part A-CD (trnL c/d), which re-extracts a
# different amplicon from the same records without re-querying NCBI. Also
# checkpoint it to scratch so a session restart before Part A-CD does not
# force repeating the query.
ncbi.trnL_raw <- ncbi.trnL
saveRDS(ncbi.trnL_raw, file.path(SCRATCH, "ncbi_trnL_raw.rds"))

ncbi.trnL <- find_primer_pair(ncbi.trnL,
                              fwd = trnL_primers[[1]],
                              rev = trnL_primers[[2]])
cat(length(ncbi.trnL), "sequences contain the trnL g/h primer pair\n")

n_before  <- length(ncbi.trnL)
ncbi.trnL <- ncbi.trnL[!grepl("UNVERIFIED", names(ncbi.trnL))]
cat("Removed", n_before - length(ncbi.trnL), "unverified sequences\n")

names(ncbi.trnL) <-
    names(ncbi.trnL) |>
    gsub(pattern = " .+$", replacement = "") |>
    gsub(pattern = "^>",   replacement = "")
head(names(ncbi.trnL))
```

## A5. Combine RefSeq and GenBank

```{r trnl_combine}
cat("RefSeq sequences: ",  length(refseq.trnL), "\n")
cat("GenBank sequences:",  length(ncbi.trnL),   "\n")
cat("Shared accessions: ", intersect(names(ncbi.trnL), names(refseq.trnL)) |> length(), "\n")
cat("RefSeq-only:       ", setdiff(names(refseq.trnL), names(ncbi.trnL))   |> length(), "\n")
cat("GenBank-only:      ", setdiff(names(ncbi.trnL),   names(refseq.trnL)) |> length(), "\n")

trnL.df <- bind_rows(
    data.frame(source = "GenBank", accession = names(ncbi.trnL),   seq = as.character(ncbi.trnL)),
    data.frame(source = "RefSeq",  accession = names(refseq.trnL), seq = as.character(refseq.trnL))
)
cat(nrow(trnL.df), "total sequences\n")
```

```{r trnl_manual_additions}
additions <- filter(edits, type == "add")

if (nrow(additions) > 0) {
    cat("Fetching", nrow(additions), "manual addition(s) from NCBI...\n")

    seqs_raw <-
        entrez_fetch(db = "nucleotide", id = additions$accession, rettype = "fasta") |>
        strsplit("\n{2,}") |>
        unlist()

    accs      <- str_extract(seqs_raw, "[^>]\\S*")
    seqs_only <- seqs_raw |>
        sub("^[^\n]*\n", "", x = _) |>
        gsub("\n", "", x = _)

    manual_seqs        <- DNAStringSet(seqs_only)
    names(manual_seqs) <- accs
    manual_seqs        <- find_primer_pair(manual_seqs,
                                           fwd = trnL_primers[[1]],
                                           rev = trnL_primers[[2]])

    trnL.df <- bind_rows(
        data.frame(seq = as.character(manual_seqs), accession = names(manual_seqs)),
        trnL.df
    )
    cat("Manual additions appended. Total:", nrow(trnL.df), "\n")
}
```

## A6. Taxonomy lookup

```{r trnl_taxonomy_lookup}
cat("Looking up taxonomy for", nrow(trnL.df), "accessions...\n")
trnL_ids <- accessionToTaxa(trnL.df$accession, SQL_PATH)
cat("Missing taxonomy IDs:", sum(is.na(trnL_ids)), "\n")

if (any(is.na(trnL_ids))) {
    cat("Accessions with missing IDs (assign manually in the next chunk):\n")
    print(trnL.df[is.na(trnL_ids), c("source", "accession")])
}
```

```{r trnl_taxonomy_missing}
# Automatically resolve missing IDs using query_ncbi_accession() for accessions
# added to NCBI after the SQL database was built
if (any(is.na(trnL_ids))) {
    missing.df <- trnL.df[is.na(trnL_ids), c("source", "accession")]
    for (i in seq_len(nrow(missing.df))) {
        idx         <- as.integer(rownames(missing.df)[i])
        acc         <- missing.df[i, 2]
        trnL_ids[idx] <- query_ncbi_accession(acc)
    }
}

cat("Remaining missing IDs:", sum(is.na(trnL_ids)), "\n")
```

```{r trnl_taxonomy_extract}
trnL_taxonomy <- extract_taxonomy(trnL_ids, SQL_PATH)
trnL.df       <- bind_cols(trnL.df, trnL_taxonomy)
head(trnL.df)
```

## A6b. Remove off-target sequences

The GenBank query can return records whose deposited taxonomy is outside the food-plant
clades entirely (bacterial or viral records co-deposited with plant material). Filter by
phylum allowlist, mirroring the 12SV5 filter at B6b.

```{r trnl_offtarget_filter}
# Cyanobacteriota is deliberately allowed: Arthrospira platensis (spirulina)
# and Nostoc flagelliforme (fat choy) are edible cyanobacteria on the food
# list with real trnL records. Keep this vector in sync with ALLOWED_PHYLA
# in code/qc_reference_build.R.
allowed_phyla_trnL <- c("Streptophyta", "Cyanobacteriota")

n_before_offtarget <- nrow(trnL.df)
trnL.df <- filter(trnL.df, phylum %in% allowed_phyla_trnL)
cat("Off-target filter: removed", n_before_offtarget - nrow(trnL.df),
    "sequences with phylum outside the food-plant allowlist\n")
cat(nrow(trnL.df), "sequences retained\n")
```

## A7. Manual curation edits

```{r trnl_curation}
# Omissions
omit    <- filter(edits, type == "omit")
trnL.df <- filter(trnL.df, !(accession %in% omit$accession & seq %in% omit$sequence))
cat("After omissions:", nrow(trnL.df), "sequences\n")

# Brassica oleracea subspecies renamings
trnL.df$varietas[trnL.df$accession == "AB213010.1"] <- "Brassica oleracea var. capitata"
trnL.df$varietas[trnL.df$accession == "AC183493.1"] <- "Brassica oleracea var. alboglabra"
trnL.df$varietas[trnL.df$accession == "LR031874.1"] <- "Brassica oleracea var. italica"
trnL.df$varietas[trnL.df$accession == "LR031875.1"] <- "Brassica oleracea var. italica"
trnL.df$varietas[trnL.df$accession == "LR031876.1"] <- "Brassica oleracea var. italica"

# Assign lowest available taxonomic rank as display name
trnL.df <-
    trnL.df |>
    MButils::lowest_level() |>
    rename(taxon = "name") |>
    select(source, accession, taxon, taxid, superkingdom:forma, seq)
```

## A8. Quality control

```{r trnl_qc}
trnL.df$N <- grepl("[AGCT]*[^AGCT]+", trnL.df$seq)
cat(sum(trnL.df$N), "sequences with degenerate nucleotides\n")

# Report whether affected taxa have a clean backup sequence
with_n <- trnL.df$taxon[trnL.df$N] |> unique()

trnL.df |>
    filter(taxon %in% with_n) |>
    group_by(taxon, N) |>
    count() |>
    ungroup() |>
    group_by(taxon) |>
    summarize(has_clean_backup = any(!N)) |>
    arrange(has_clean_backup)
```

```{r trnl_remove_degenerate}
trnL.df <- filter(trnL.df, !grepl("[AGCT]*[^AGCT]+", seq))
cat(nrow(trnL.df), "sequences after removing degenerate sequences\n")
```

## A9. Standardise orientation

```{r trnl_orientation}
fwd_err <- floor(0.2 * length(trnLG))
rev_err <- floor(0.2 * length(trnLH))
ref     <- DNAStringSet(trnL.df$seq)
names(ref) <- paste(trnL.df$accession, trnL.df$taxon)

fwd_matches <-
    vmatchPattern(trnLG, ref, max.mismatch = fwd_err, fixed = TRUE) |>
    as.data.frame() |>
    filter(start <= 1) |>
    mutate(type = "forward") |>
    select(group, type)

rev_matches <-
    vmatchPattern(trnLH, ref, max.mismatch = rev_err, fixed = TRUE) |>
    as.data.frame() |>
    filter(start <= 1) |>
    mutate(type = "reverse") |>
    select(group, type)

# Keep the orientation table: the summary below needs `type` after it has been
# dropped from trnL.df.
orientation <-
    bind_rows(fwd_matches, rev_matches) |>
    arrange(group)

trnL.df <-
    orientation |>
    bind_cols(trnL.df) |>
    mutate(seq = ifelse(type == "reverse",
                        DNAStringSet(seq) |> reverseComplement() |> as.character(),
                        seq)) |>
    select(-c(type, group))

count(orientation, type)
```

## A10. Deduplicate, add controls, and save

```{r trnl_dedup}
trnL.df <-
    trnL.df |>
    group_by(superkingdom, phylum, class, order, family,
             genus, species, subspecies, varietas, forma, taxon, seq) |>
    arrange(desc(source), accession) |>
    summarize(accession = first(accession), .groups = "drop") |>
    arrange(taxon, accession)

cat(nrow(trnL.df), "sequences after deduplication\n")
cat(n_distinct(trnL.df$taxon), "unique taxa\n")
```

Append the positive-control spike-ins defined in `data/inputs/controls.csv`. These are
laboratory constructs, not NCBI records, so they never come out of the RefSeq/GenBank
queries above and must be added here or they are silently absent from every rebuild.

The lab convention repeats the control label across `superkingdom:species` rather than
placing it at a single rank. That matters for two reasons: DADA2 then reports the label at
every level and it survives last-common-ancestor condensing, and the genus field stays
non-`NA`, which keeps the record from being stripped by downstream genus-NA filtering.

```{r trnl_controls}
controls.trnL <-
    read_csv(file.path(REPO_DIR, "data", "inputs", "controls.csv"),
             show_col_types = FALSE) |>
    filter(marker == "trnL")

if (nrow(controls.trnL) > 0) {
    control_rows <-
        controls.trnL |>
        # Two kinds of control: synthetic constructs (empty lineage) get the
        # label repeated across superkingdom:species, per the lab convention;
        # real-species controls carry their true lineage from the lineage
        # column so DADA2 places them correctly.
        mutate(lineage = ifelse(is.na(lineage) | lineage == "",
                                sapply(label, function(l)
                                    paste(c(rep(l, 7), rep("NA", 3)),
                                          collapse = ";")),
                                lineage)) |>
        separate(lineage,
                 into = c("superkingdom", "phylum", "class", "order", "family",
                          "genus", "species", "subspecies", "varietas", "forma"),
                 sep = ";") |>
        mutate(across(superkingdom:forma, \(x) na_if(x, "NA"))) |>
        transmute(superkingdom, phylum, class, order, family, genus, species,
                  subspecies, varietas, forma,
                  taxon = coalesce(species, label), seq = sequence, accession) |>
        # guard against double-adding if this chunk is re-run
        filter(!(seq %in% trnL.df$seq))

    trnL.df <- bind_rows(trnL.df, control_rows)
    cat("Added", nrow(control_rows), "positive-control sequence(s):",
        paste(control_rows$taxon, collapse = ", "), "\n")
} else {
    warning("No trnL controls found in controls.csv - reference will have no positive control.")
}

cat(nrow(trnL.df), "sequences total (including controls)\n")
```

```{r trnl_save}
# DADA2: sequence file
trnL_out <- DNAStringSet(trnL.df$seq)
names(trnL_out) <- paste(trnL.df$accession, trnL.df$taxon)
writeXStringSet(trnL_out, file.path(TRNL_DADA2, "trnLGH.fasta"))
cat("Saved trnLGH.fasta\n")

# DADA2: taxonomy file
trnL.df_tax     <- trnL.df |> unite(col = "lineage", superkingdom:forma, sep = ";")
names(trnL_out) <- trnL.df_tax$lineage
writeXStringSet(trnL_out, file.path(TRNL_DADA2, "trnLGH_taxonomy.fasta"))
cat("Saved trnLGH_taxonomy.fasta\n")

# QIIME2
trnL.df_tax     <- mutate(trnL.df_tax, acc = make.unique(accession, sep = "_"))
names(trnL_out) <- trnL.df_tax$acc
writeXStringSet(trnL_out, file.path(TRNL_QIIME2, "trnL-sequences.fasta"))
trnL.df_tax |>
    select(`Feature ID` = acc, Taxon = lineage) |>
    write_delim(file.path(TRNL_QIIME2, "trnL-taxonomy.tsv"), delim = "\t")
cat("Saved QIIME2 files\n")

cat("\n=== trnL complete:", nrow(trnL.df), "sequences,",
    n_distinct(trnL.df$taxon), "taxa ===\n")
```

---

# Part A-CD — trnL c/d Reference Database

Primers: trnL-c (`CGAAATCGGTAGACGCTACG`) / trnL-d (`GGGGATAGAGGGACTTGAAC`) — the full
trnL(UAA) intron, a longer amplicon (~300–600 bp) with higher taxonomic resolution than
the g/h P6 loop.

This part reuses Part A's inputs directly — the RefSeq plastome set from A2 and the
untrimmed GenBank pull stashed in A4 (`ncbi.trnL_raw`) — so it adds **no** NCBI queries.
It must run after Part A (it also reuses `plants` and `allowed_phyla_trnL`).

Extraction is **alignment-augmented**, in three tiers:

1. **Plastome introns** — extracted by primer sites from the RefSeq plastomes (complete
   genomes always carry the sites). These become reference records AND the alignment panel.
2. **GenBank primer route** — records that carry both priming sites are extracted directly.
3. **GenBank alignment rescue** — records without primer sites (most long trnL deposits are
   primer-trimmed introns) are locally aligned against family-matched plastome introns and
   the homologous region is extracted.

All records are stored **primer-trimmed and c→d oriented**. That matches how reads are
processed (primers removed before denoising) and matches the format of the 2025 trnLCD
compilation.

> **Method basis (validated 2026-08-13):** primer-only extraction reaches ~55–60% of the
> taxa the g/h reference covers, because most long GenBank trnL deposits carry neither c/d
> site (primer-trimmed introns and trnL-F spacer records). Alignment localized the intron
> in 99/100 sampled primer-less taxa, and the 2025 `trnLCD.fasta` in
> `data/outputs/dada2-compatible/miscellaneous/` (Ashish Subramanian; demonstrated to be
> annotation/alignment-based, not primer-based) reached ~90% the same way. Expected
> coverage here: ~90–95% of g/h taxa. The CD8 anchor check guards extraction quality.

## CD1. Primers and alignment setup

```{r trnlcd_inputs}
trnLC <- DNAString("CGAAATCGGTAGACGCTACG")
trnLD <- DNAString("GGGGATAGAGGGACTTGAAC")
trnLCD_primers <- list(trnLC, trnLD)

# Fresh-session recovery: the untrimmed A4 pull is checkpointed to scratch.
# (In a fresh session, first rerun Configuration, Libraries, A1, A2, and
# define allowed_phyla_trnL as in A6b.)
if (!exists("ncbi.trnL_raw"))
    ncbi.trnL_raw <- readRDS(file.path(SCRATCH, "ncbi_trnL_raw.rds"))

# pairwiseAlignment machinery lives in pwalign on Bioconductor >= 3.19 and in
# Biostrings itself before that; bind whichever this installation provides.
if (requireNamespace("pwalign", quietly = TRUE)) {
    pa_align   <- pwalign::pairwiseAlignment
    pa_nsm     <- pwalign::nucleotideSubstitutionMatrix
    pa_pid     <- pwalign::pid
    pa_pattern <- pwalign::pattern
} else {
    pa_align   <- Biostrings::pairwiseAlignment
    pa_nsm     <- Biostrings::nucleotideSubstitutionMatrix
    pa_pid     <- Biostrings::pid
    pa_pattern <- Biostrings::pattern
}

CD_SUB_MAT  <- pa_nsm(match = 1, mismatch = -1, baseOnly = FALSE)
CD_MIN_PID  <- 60     # min % identity of the best local alignment
CD_MIN_SPAN <- 150    # min bp extracted (2025 compilation minimum was 152)
CD_CAND_MIN <- 250    # candidate record length window for alignment rescue
CD_CAND_MAX <- 20000

# Orient an amplicon so trnL-c is at its 5' end (whichever orientation places
# it there with fewer edits), then trim both 20 bp primers off.
orient_and_trim_cd <- function(amps) {
    fwd_mm <- vapply(seq_along(amps), function(i)
        neditStartingAt(trnLC, amps[[i]], starting.at = 1), 0L)
    rev_mm <- vapply(seq_along(amps), function(i)
        neditStartingAt(trnLC, reverseComplement(amps[[i]]), starting.at = 1), 0L)
    flip <- rev_mm < fwd_mm
    amps[flip] <- reverseComplement(amps[flip])
    subseq(amps, 21L, width(amps) - 20L)
}
```

## CD2. Plastome intron panel (reference records + alignment panel)

```{r trnlcd_refseq_panel}
refseq.ampCD <- find_primer_pair(plastid_species,
                                 fwd = trnLCD_primers[[1]],
                                 rev = trnLCD_primers[[2]])
cat(length(refseq.ampCD), "plastome sequences contain the trnL c/d primer pair\n")

plants.cd.i  <- lapply(plants, grep, x = names(refseq.ampCD)) |> unlist() |> unique()
refseq.ampCD <- refseq.ampCD[plants.cd.i]
refseq.ampCD <- refseq.ampCD[width(refseq.ampCD) >= 100]
cat(length(refseq.ampCD), "food plant plastome amplicons\n")

refseq.intronCD <- orient_and_trim_cd(refseq.ampCD)
names(refseq.intronCD) <- gsub(" .*$", "", names(refseq.intronCD))

# Family index for family-matched alignment in CD3b
panel_ids <- accessionToTaxa(names(refseq.intronCD), SQL_PATH)
panel_fam <- getTaxonomy(panel_ids, SQL_PATH, desiredRanks = "family")[, "family"]
cat(length(refseq.intronCD), "plastome introns spanning",
    n_distinct(na.omit(panel_fam)), "families\n")
```

## CD3. GenBank: primer route

```{r trnlcd_ncbi_pcr}
ncbi.ampCD <- find_primer_pair(ncbi.trnL_raw,
                               fwd = trnLCD_primers[[1]],
                               rev = trnLCD_primers[[2]])
cat(length(ncbi.ampCD), "GenBank records contain the trnL c/d primer pair\n")

ncbi.ampCD <- ncbi.ampCD[!grepl("UNVERIFIED", names(ncbi.ampCD))]
ncbi.ampCD <- ncbi.ampCD[width(ncbi.ampCD) >= 100]

ncbi.intronCD <- orient_and_trim_cd(ncbi.ampCD)
names(ncbi.intronCD) <-
    names(ncbi.intronCD) |>
    gsub(pattern = " .+$", replacement = "") |>
    gsub(pattern = "^>",   replacement = "")
cat(length(ncbi.intronCD), "primer-extracted GenBank introns\n")
```

## CD3b. GenBank: alignment rescue

Records without primer sites but long enough to hold the intron are aligned against
family-matched plastome introns (falling back to a taxonomically spread set when the
family is unknown or unrepresented), and the homologous region is extracted. Identical
sequences are aligned once and the result mapped back to every accession sharing them.

```{r trnlcd_align_rescue}
cand <- ncbi.trnL_raw[!grepl("UNVERIFIED", names(ncbi.trnL_raw))]
names(cand) <-
    names(cand) |>
    gsub(pattern = " .+$", replacement = "") |>
    gsub(pattern = "^>",   replacement = "")
cand <- cand[width(cand) >= CD_CAND_MIN & width(cand) <= CD_CAND_MAX]
cand <- cand[!(names(cand) %in% names(ncbi.intronCD))]
cat(length(cand), "candidate records for alignment rescue\n")

cand_chr <- as.character(cand)
uniq_i   <- which(!duplicated(cand_chr))
uniq     <- cand[uniq_i]
uniq_ids <- accessionToTaxa(names(uniq), SQL_PATH)
uniq_fam <- getTaxonomy(uniq_ids, SQL_PATH, desiredRanks = "family")[, "family"]
cat(length(uniq), "unique candidate sequences\n")

fallback_refs <- refseq.intronCD[head(which(!duplicated(panel_fam) & !is.na(panel_fam)), 8)]

align_extract_cd <- function(s, fam) {
    refs <- fallback_refs
    if (!is.na(fam)) {
        fi <- which(!is.na(panel_fam) & panel_fam == fam)
        if (length(fi) > 0) refs <- refseq.intronCD[head(fi, 3)]
    }
    # orient once against the first reference
    a_f <- pa_align(s, refs[[1]], type = "local", substitutionMatrix = CD_SUB_MAT,
                    gapOpening = 2, gapExtension = 0.5)
    a_r <- pa_align(reverseComplement(s), refs[[1]], type = "local",
                    substitutionMatrix = CD_SUB_MAT, gapOpening = 2, gapExtension = 0.5)
    if (score(a_r) > score(a_f)) s <- reverseComplement(s)
    best <- NULL
    for (k in seq_along(refs)) {
        a  <- pa_align(s, refs[[k]], type = "local", substitutionMatrix = CD_SUB_MAT,
                       gapOpening = 2, gapExtension = 0.5)
        st <- start(pa_pattern(a)); en <- end(pa_pattern(a))
        if (is.null(best) || (en - st + 1) > best$span)
            best <- list(span = en - st + 1, pid = pa_pid(a), st = st, en = en)
    }
    if (best$span >= CD_MIN_SPAN && best$pid >= CD_MIN_PID)
        as.character(subseq(s, best$st, best$en))
    else NA_character_
}

uniq_intron <- rep(NA_character_, length(uniq))
for (i in seq_along(uniq)) {
    uniq_intron[i] <- align_extract_cd(uniq[[i]], uniq_fam[i])
    if (i %% 1000 == 0) cat("  aligned", i, "of", length(uniq), "candidates\n")
}
cat(sum(!is.na(uniq_intron)), "of", length(uniq),
    "unique candidates yielded an intron\n")

map          <- match(cand_chr, cand_chr[uniq_i])
aligned_seqs <- uniq_intron[map]
keep         <- !is.na(aligned_seqs)
aligned.df   <- data.frame(source    = "GenBank-aligned",
                           accession = names(cand)[keep],
                           seq       = aligned_seqs[keep])
cat(nrow(aligned.df), "alignment-rescued records\n")
```

## CD4. Combine all three tiers

```{r trnlcd_combine}
trnLCD.df <- bind_rows(
    data.frame(source = "GenBank", accession = names(ncbi.intronCD),   seq = as.character(ncbi.intronCD)),
    aligned.df,
    data.frame(source = "RefSeq",  accession = names(refseq.intronCD), seq = as.character(refseq.intronCD))
)
cat(nrow(trnLCD.df), "total sequences\n")
print(table(trnLCD.df$source))
```

## CD5. Taxonomy lookup

```{r trnlcd_taxonomy}
cat("Looking up taxonomy for", nrow(trnLCD.df), "accessions...\n")
trnLCD_ids <- accessionToTaxa(trnLCD.df$accession, SQL_PATH)
cat("Missing taxonomy IDs:", sum(is.na(trnLCD_ids)), "\n")

if (any(is.na(trnLCD_ids))) {
    missing.df <- trnLCD.df[is.na(trnLCD_ids), c("source", "accession")]
    for (i in seq_len(nrow(missing.df))) {
        idx             <- as.integer(rownames(missing.df)[i])
        acc             <- missing.df[i, 2]
        trnLCD_ids[idx] <- query_ncbi_accession(acc)
    }
    cat("Remaining missing IDs:", sum(is.na(trnLCD_ids)), "\n")
}

trnLCD.df <- bind_cols(trnLCD.df, extract_taxonomy(trnLCD_ids, SQL_PATH))
```

## CD6. Remove off-target sequences

```{r trnlcd_offtarget_filter}
# Same allowlist as A6b (allowed_phyla_trnL is defined there); keep in sync
# with ALLOWED_PHYLA in code/qc_reference_build.R.
n_before_offtarget <- nrow(trnLCD.df)
trnLCD.df <- filter(trnLCD.df, phylum %in% allowed_phyla_trnL)
cat("Off-target filter: removed", n_before_offtarget - nrow(trnLCD.df),
    "sequences with phylum outside the food-plant allowlist\n")
cat(nrow(trnLCD.df), "sequences retained\n")
```

## CD7. Naming

```{r trnlcd_naming}
# Brassica oleracea subspecies renamings — same source records as A7, applied
# when those accessions yield a c/d amplicon. (A7's omission edits are keyed to
# g/h amplicon sequences and cannot match here, so they are not repeated.)
trnLCD.df$varietas[trnLCD.df$accession == "AB213010.1"] <- "Brassica oleracea var. capitata"
trnLCD.df$varietas[trnLCD.df$accession == "AC183493.1"] <- "Brassica oleracea var. alboglabra"
trnLCD.df$varietas[trnLCD.df$accession == "LR031874.1"] <- "Brassica oleracea var. italica"
trnLCD.df$varietas[trnLCD.df$accession == "LR031875.1"] <- "Brassica oleracea var. italica"
trnLCD.df$varietas[trnLCD.df$accession == "LR031876.1"] <- "Brassica oleracea var. italica"

# Assign lowest available taxonomic rank as display name
trnLCD.df <-
    trnLCD.df |>
    MButils::lowest_level() |>
    rename(taxon = "name") |>
    select(source, accession, taxon, taxid, superkingdom:forma, seq)
```

## CD8. Quality control

```{r trnlcd_remove_degenerate}
cat(sum(grepl("[AGCT]*[^AGCT]+", trnLCD.df$seq)),
    "sequences with degenerate nucleotides - removing\n")
trnLCD.df <- filter(trnLCD.df, !grepl("[AGCT]*[^AGCT]+", seq))
cat(nrow(trnLCD.df), "sequences remaining\n")
```

## CD9. Extraction sanity check

Orientation is standardised at extraction time (all three tiers emit c→d oriented,
primer-trimmed introns), so no orientation pass is needed. Instead, verify extraction
quality: the trnL-g primer anneals **inside** the intron, so nearly every true intron
must contain a g-like site. A low anchor rate means the alignment rescue is extracting
the wrong region.

```{r trnlcd_anchor_check}
g_anchor <- vcountPattern(DNAString("GGGCAATCCTGAGCCAA"),
                          DNAStringSet(trnLCD.df$seq), max.mismatch = 3) > 0
cat(sprintf("%.1f%% of records contain the internal trnL-g anchor\n",
            100 * mean(g_anchor)))
cat("anchor rate by source tier:\n")
print(round(100 * tapply(g_anchor, trnLCD.df$source, mean), 1))
if (mean(g_anchor) < 0.80)
    warning("Low trnL-g anchor rate - review CD3b extraction thresholds before shipping.")
```

## CD10. Deduplicate, add controls, and save

```{r trnlcd_dedup}
trnLCD.df <-
    trnLCD.df |>
    group_by(superkingdom, phylum, class, order, family,
             genus, species, subspecies, varietas, forma, taxon, seq) |>
    arrange(desc(source), accession) |>
    summarize(accession = first(accession), .groups = "drop") |>
    arrange(taxon, accession)

cat(nrow(trnLCD.df), "sequences after deduplication\n")
cat(n_distinct(trnLCD.df$taxon), "unique taxa\n")
```

```{r trnlcd_controls}
controls.trnLCD <-
    read_csv(file.path(REPO_DIR, "data", "inputs", "controls.csv"),
             show_col_types = FALSE) |>
    filter(marker == "trnLCD")

if (nrow(controls.trnLCD) > 0) {
    control_rows <-
        controls.trnLCD |>
        mutate(lineage = ifelse(is.na(lineage) | lineage == "",
                                sapply(label, function(l)
                                    paste(c(rep(l, 7), rep("NA", 3)),
                                          collapse = ";")),
                                lineage)) |>
        separate(lineage,
                 into = c("superkingdom", "phylum", "class", "order", "family",
                          "genus", "species", "subspecies", "varietas", "forma"),
                 sep = ";") |>
        mutate(across(superkingdom:forma, \(x) na_if(x, "NA"))) |>
        transmute(superkingdom, phylum, class, order, family, genus, species,
                  subspecies, varietas, forma,
                  taxon = coalesce(species, label), seq = sequence, accession) |>
        filter(!(seq %in% trnLCD.df$seq))

    trnLCD.df <- bind_rows(trnLCD.df, control_rows)
    cat("Added", nrow(control_rows), "positive-control sequence(s):",
        paste(control_rows$taxon, collapse = ", "), "\n")
} else {
    # No c/d control is defined yet: the 50 bp synthetic construct carries no
    # c/d priming sites. This is a deliberate gap, not an error — add a row
    # with marker "trnLCD" to controls.csv once the wet lab settles on one.
    cat("No trnLCD controls defined in controls.csv (expected for now).\n")
}

cat(nrow(trnLCD.df), "sequences total\n")
```

```{r trnlcd_save}
# DADA2: sequence file
trnLCD_out <- DNAStringSet(trnLCD.df$seq)
names(trnLCD_out) <- paste(trnLCD.df$accession, trnLCD.df$taxon)
writeXStringSet(trnLCD_out, file.path(TRNLCD_DADA2, "trnLCD.fasta"))
cat("Saved trnLCD.fasta\n")

# DADA2: taxonomy file
trnLCD.df_tax     <- trnLCD.df |> unite(col = "lineage", superkingdom:forma, sep = ";")
names(trnLCD_out) <- trnLCD.df_tax$lineage
writeXStringSet(trnLCD_out, file.path(TRNLCD_DADA2, "trnLCD_taxonomy.fasta"))
cat("Saved trnLCD_taxonomy.fasta\n")

# QIIME2
trnLCD.df_tax     <- mutate(trnLCD.df_tax, acc = make.unique(accession, sep = "_"))
names(trnLCD_out) <- trnLCD.df_tax$acc
writeXStringSet(trnLCD_out, file.path(TRNLCD_QIIME2, "trnLCD-sequences.fasta"))
trnLCD.df_tax |>
    select(`Feature ID` = acc, Taxon = lineage) |>
    write_delim(file.path(TRNLCD_QIIME2, "trnLCD-taxonomy.tsv"), delim = "\t")
cat("Saved QIIME2 files\n")

cat("\n=== trnL c/d complete:", nrow(trnLCD.df), "sequences,",
    n_distinct(trnLCD.df$taxon), "taxa ===\n")
```

---

# Part B — 12SV5 Reference Database

Primers: V5F (`TAGAACAGGCTCCTCTAG`) / V5R (`TTAGATACCCCACTATGC`)
Source organisms: curated list of dietary vertebrates (extend `animals` below as needed)

> **Note:** The original `acc_to_dada()` function in the repo has incomplete code
> and a hardcoded legacy path. This pipeline uses the same taxonomizr-based approach
> as Part A instead, which is more robust and uses the same SQL database.

## B1. Load inputs

```{r s12v5_inputs}
V5F <- DNAString("TAGAACAGGCTCCTCTAG")
V5R <- DNAString("TTAGATACCCCACTATGC")
s12v5_primers <- list(V5F, V5R)

# Read dietary animal list from the same curated CSV used for plants
animals <-
    file.path(REPO_DIR, "data", "inputs", "human-foods.csv") |>
    read.csv(stringsAsFactors = FALSE) |>
    filter(category == "animal") |>
    pull(scientific_name)

cat(length(animals), "dietary animal species\n")
head(animals)
tail(animals)
```

## B2. Filter RefSeq mitochondrial to one sequence per species

```{r s12v5_refseq_filter}
# "_acc" filename: names are "<accession> <Genus species>". Do NOT fall back to
# an old refseq_mito_species.fasta cache — its names are species-only, which is
# the bug this replaces (see below).
mito_species_path <- file.path(MITO_DIR, "refseq_mito_species_acc.fasta")

if (!file.exists(mito_species_path)) {
    cat("Filtering RefSeq mitochondrial to one sequence per species...\n")
    mito <- readDNAStringSet(file.path(MITO_DIR, "refseq_mito_all.fasta"))
    cat(length(mito), "total sequences\n")

    # Strip 'mitochondrion' suffix for cleaner species name extraction
    names(mito) <- names(mito) |>
        strsplit(" mitochondrion") |>
        lapply(`[[`, 1) |>
        unlist()

    # The accession MUST stay in the name: B3 keeps the first token as the
    # record's accession, and taxonomy lookup depends on it. The May 2026 run
    # named these by species only, which put a bare GENUS in the accession
    # column of all 260 RefSeq-sourced records and sent their taxonomy lookups
    # through the free-text NCBI fallback — 39 came back cross-genus mislabels
    # (Esox->Salmo salar etc.).
    accession    <- sub("^(\\S+).*", "\\1", names(mito))
    species      <- sub("^\\S+\\s+(\\S+\\s+\\S+).*", "\\1", names(mito))
    mito_species <- mito[!duplicated(species)]
    names(mito_species) <- paste(accession, species)[!duplicated(species)]

    writeXStringSet(mito_species, mito_species_path)
    cat(length(mito_species), "species retained\n")
} else {
    mito_species <- readDNAStringSet(mito_species_path)
    cat(length(mito_species), "species loaded from existing file\n")
}

mito_species
```

## B3. RefSeq: in silico PCR and subset to dietary animals

```{r s12v5_refseq_pcr}
refseq.12SV5 <- find_primer_pair(clean(mito_species),
                                 fwd = s12v5_primers[[1]],
                                 rev = s12v5_primers[[2]])
cat(length(refseq.12SV5), "sequences contain the 12SV5 primer pair\n")

animals.i    <- lapply(animals, grep, x = names(refseq.12SV5)) |> unlist()
refseq.12SV5 <- refseq.12SV5[animals.i]
cat(length(refseq.12SV5), "dietary animal 12SV5 sequences from RefSeq\n")

# Report any animals not found in RefSeq
found   <- sub("\\S+\\s", "", names(refseq.12SV5))
missing <- animals[sapply(animals, function(x) length(grep(x, found))) == 0]
if (length(missing) > 0)
    cat("Not found in RefSeq (will rely on GenBank):", paste(missing, collapse = ", "), "\n")

names(refseq.12SV5) <- gsub(" .*$", "", names(refseq.12SV5))
head(names(refseq.12SV5))
```

## Save B3 results for SLURM job

```{r s12v5_save_b3}
saveRDS(refseq.12SV5, file.path(SCRATCH, "refseq_12sv5.rds"))
saveRDS(animals,      file.path(SCRATCH, "animals_12sv5.rds"))
cat("Saved refseq.12SV5 and animals to", SCRATCH, "\n")
```

## Submit B4–B6 as a SLURM job (recommended)

B4 (NCBI GenBank query) and the missing-ID loop in B6 both make thousands of
API calls and can take several hours. Run them as a SLURM batch job to avoid
session timeouts. Once complete, load the results with the chunk below and
proceed to B7.

```{r submit_b4b6}
# ── Standalone R script ────────────────────────────────────────────────────
b4b6_r_path <- file.path(SCRATCH, "run_b4_b6.R")

r_lines <- c(
    '# run_b4_b6.R — 12SV5 GenBank query and taxonomy lookup (B4-B6)',
    '# Standalone script; run as a SLURM job after saving B3 results from RStudio.',
    '',
    '# ── Configuration ────────────────────────────────────────────────────',
    paste0('SCRATCH  <- "', SCRATCH,  '"'),
    paste0('REPO_DIR <- "', REPO_DIR, '"'),
    paste0('SQL_PATH <- "', SQL_PATH, '"'),
    '',
    '# ── NCBI API key ─────────────────────────────────────────────────────',
    '# Set the ENTREZ_KEY environment variable in your ~/.bashrc to use your key.',
    '# e.g.  export ENTREZ_KEY="your_key_here"',
    'api_key <- Sys.getenv("ENTREZ_KEY", unset = "")',
    'if (nchar(api_key) > 0) {',
    '    rentrez::set_entrez_key(api_key)',
    '    cat("NCBI API key set.\\n")',
    '} else {',
    '    cat("No NCBI API key found — running at unauthenticated rate limit.\\n")',
    '}',
    '',
    '# ── Libraries ─────────────────────────────────────────────────────────',
    'library(Biostrings)',
    'library(taxonomizr)',
    'library(tidyverse)',
    'library(rentrez)',
    '',
    '# ── Source functions ──────────────────────────────────────────────────',
    'source(file.path(REPO_DIR, "code", "functions", "find_primer_pair.R"))',
    'source(file.path(REPO_DIR, "code", "functions", "query_ncbi.R"))',
    'source(file.path(REPO_DIR, "code", "functions", "query_ncbi_accession.R"))',
    '',
    '# ── Inline MButils functions (avoids system-library dependency) ───────',
    '# clean(): removes sequences containing non-ACGT characters (Ns, degenerate',
    '# bases) before primer matching, as required by find_primer_pair()',
    'clean <- function(seqs) {',
    '    seqs[!grepl("[^ACGTacgt]", as.character(seqs))]',
    '}',
    '',
    '# lowest_level(): assigns the lowest available taxonomic rank as display name',
    'lowest_level <- function(df) {',
    '    tax_ranks <- c("forma", "varietas", "subspecies", "species",',
    '                   "genus", "family", "order", "class", "phylum", "superkingdom")',
    '    present <- tax_ranks[tax_ranks %in% names(df)]',
    '    df$name <- apply(df[, present, drop = FALSE], 1, function(row) {',
    '        vals <- row[!is.na(row) & nchar(trimws(as.character(row))) > 0]',
    '        if (length(vals) > 0) vals[[1]] else NA_character_',
    '    })',
    '    df',
    '}',
    '',
    '# ── Primers ───────────────────────────────────────────────────────────',
    'V5F <- DNAString("TAGAACAGGCTCCTCTAG")',
    'V5R <- DNAString("TTAGATACCCCACTATGC")',
    's12v5_primers <- list(V5F, V5R)',
    '',
    '# ── Shared helper: accession IDs to full taxonomy data frame ──────────',
    '# NCBI reclassified Eukaryota/Bacteria/Archaea from rank "superkingdom" to',
    '# rank "domain" (taxonomizr >= 0.11 follows suit). Capture both names and',
    '# merge into "superkingdom" below, or a current taxdump yields 100% NA there.',
    'extract_taxonomy <- function(ids, sql_path) {',
    '    vars <- c("superkingdom", "domain", "phylum", "class", "order",',
    '              "family", "genus", "species", "subspecies", "varietas", "forma")',
    '    taxonomy.raw <- getRawTaxonomy(ids, sql_path)',
    '    taxonomy <- as.data.frame(',
    '        setNames(lapply(vars, function(v) character(0)), vars),',
    '        stringsAsFactors = FALSE',
    '    )',
    '    for (i in seq_along(taxonomy.raw)) {',
    '        row.i  <- taxonomy.raw[[i]] |> t() |> data.frame()',
    '        shared <- intersect(vars, names(row.i))',
    '        row.i  <- select(row.i, one_of(shared))',
    '        taxonomy <- bind_rows(taxonomy, row.i)',
    '    }',
    '    taxonomy$superkingdom <- coalesce(taxonomy$superkingdom, taxonomy$domain)',
    '    taxonomy$domain <- NULL',
    '    taxonomy$taxid <- names(taxonomy.raw) |> trimws() |> as.integer()',
    '    select(taxonomy, taxid, everything())',
    '}',
    '',
    '# ── Load B3 results ───────────────────────────────────────────────────',
    'cat("\\n=== Loading B3 results ===\\n")',
    'refseq.12SV5 <- readRDS(file.path(SCRATCH, "refseq_12sv5.rds"))',
    'animals      <- readRDS(file.path(SCRATCH, "animals_12sv5.rds"))',
    'cat(length(refseq.12SV5), "RefSeq sequences loaded\\n")',
    'cat(length(animals), "dietary animal species loaded\\n")',
    '',
    '# ── B4. NCBI GenBank query ────────────────────────────────────────────',
    'cat("\\n=== B4. NCBI GenBank query ===\\n")',
    'ncbi.12SV5 <- query_ncbi(marker = "12S", organisms = animals)',
    'cat(length(ncbi.12SV5), "sequences returned from NCBI\\n")',
    '',
    'ncbi.12SV5 <- find_primer_pair(clean(ncbi.12SV5),',
    '                               fwd = s12v5_primers[[1]],',
    '                               rev = s12v5_primers[[2]])',
    'cat(length(ncbi.12SV5), "sequences contain the 12SV5 primer pair\\n")',
    '',
    'n_before   <- length(ncbi.12SV5)',
    'ncbi.12SV5 <- ncbi.12SV5[!grepl("UNVERIFIED", names(ncbi.12SV5))]',
    'cat("Removed", n_before - length(ncbi.12SV5), "unverified sequences\\n")',
    '',
    'names(ncbi.12SV5) <-',
    '    names(ncbi.12SV5) |>',
    '    gsub(pattern = " .+$", replacement = "") |>',
    '    gsub(pattern = "^>",   replacement = "")',
    '',
    '# ── B5. Combine RefSeq and GenBank ───────────────────────────────────',
    'cat("\\n=== B5. Combine RefSeq and GenBank ===\\n")',
    'cat("RefSeq sequences: ",  length(refseq.12SV5), "\\n")',
    'cat("GenBank sequences:",  length(ncbi.12SV5),   "\\n")',
    's12v5.df <- bind_rows(',
    '    data.frame(source = "GenBank", accession = names(ncbi.12SV5),   seq = as.character(ncbi.12SV5)),',
    '    data.frame(source = "RefSeq",  accession = names(refseq.12SV5), seq = as.character(refseq.12SV5))',
    ')',
    'cat(nrow(s12v5.df), "total sequences\\n")',
    '',
    '# ── B6. Taxonomy lookup ───────────────────────────────────────────────',
    'cat("\\n=== B6. Taxonomy lookup ===\\n")',
    'cat("Looking up taxonomy for", nrow(s12v5.df), "accessions...\\n")',
    's12v5_ids <- accessionToTaxa(s12v5.df$accession, SQL_PATH)',
    'cat("Missing taxonomy IDs:", sum(is.na(s12v5_ids)), "\\n")',
    '',
    'checkpoint_path <- file.path(SCRATCH, "s12v5_ids_checkpoint.rds")',
    '',
    'if (any(is.na(s12v5_ids))) {',
    '    # Resume from checkpoint if one exists (e.g. after a time-limit cancel)',
    '    if (file.exists(checkpoint_path)) {',
    '        s12v5_ids <- readRDS(checkpoint_path)',
    '        cat("Checkpoint loaded:", sum(!is.na(s12v5_ids)), "IDs already resolved\\n")',
    '    }',
    '    missing.df <- s12v5.df[is.na(s12v5_ids), c("source", "accession")]',
    '    cat("Resolving", nrow(missing.df), "missing IDs via NCBI API...\\n")',
    '    for (i in seq_len(nrow(missing.df))) {',
    '        idx            <- as.integer(rownames(missing.df)[i])',
    '        acc            <- missing.df[i, 2]',
    '        s12v5_ids[idx] <- tryCatch(query_ncbi_accession(acc),',
    '                                   error = function(e) NA_integer_)',
    '        if (i %% 100 == 0) {',
    '            cat("  Resolved", i, "of", nrow(missing.df), "missing IDs\\n")',
    '            saveRDS(s12v5_ids, checkpoint_path)',
    '        }',
    '    }',
    '    cat("Remaining missing IDs:", sum(is.na(s12v5_ids)), "\\n")',
    '    file.remove(checkpoint_path)   # clean up on successful completion',
    '}',
    '',
    's12v5_taxonomy <- extract_taxonomy(s12v5_ids, SQL_PATH)',
    's12v5.df       <- bind_cols(s12v5.df, s12v5_taxonomy)',
    '',
    's12v5.df <-',
    '    s12v5.df |>',
    '    lowest_level() |>',
    '    rename(taxon = "name") |>',
    '    select(source, accession, taxon, taxid, superkingdom:forma, seq)',
    '',
    '# ── Save results ──────────────────────────────────────────────────────',
    'cat("\\n=== Saving results ===\\n")',
    'saveRDS(s12v5.df,  file.path(SCRATCH, "s12v5_df_b6.rds"))',
    'saveRDS(s12v5_ids, file.path(SCRATCH, "s12v5_ids_b6.rds"))',
    'cat("Saved s12v5_df_b6.rds and s12v5_ids_b6.rds to", SCRATCH, "\\n")',
    'cat("\\n=== B4-B6 complete ===\\n")'
)

writeLines(r_lines, b4b6_r_path)
cat("R script written to:", b4b6_r_path, "\n")

# ── SLURM bash script ──────────────────────────────────────────────────────
b4b6_slurm <- paste0(
"#!/bin/bash\n",
"#SBATCH --job-name=12sv5_b4b6\n",
"#SBATCH --partition=", PARTITION, "\n",
"#SBATCH --time=24:00:00\n",
"#SBATCH --mem=128G\n",
"#SBATCH --cpus-per-task=2\n",
"#SBATCH --output=", SCRATCH, "/12sv5_b4b6.log\n",
"\n",
"module load ", R_MODULE, "\n",
"\n",
"# To use your NCBI API key, uncomment and fill in the line below:\n",
"# export ENTREZ_KEY='your_key_here'\n",
"\n",
"Rscript ", b4b6_r_path, "\n",
"echo 'B4-B6 job finished.'\n"
)

b4b6_sh <- file.path(SCRATCH, "submit_b4_b6.sh")
writeLines(b4b6_slurm, b4b6_sh)

cat("SLURM script written to:", b4b6_sh, "\n\n")
cat("Submit from the OOD terminal with:\n")
cat("  sbatch", b4b6_sh, "\n\n")
cat("Monitor progress with:\n")
cat("  tail -f", file.path(SCRATCH, "12sv5_b4b6.log"), "\n")
```

## Monitor B4–B6 job

Re-run this chunk to check job progress.

```{r monitor_b4b6}
log_path <- file.path(SCRATCH, "12sv5_b4b6.log")

if (file.exists(log_path)) {
    lines     <- readLines(log_path)
    completed  <- any(grepl("B4-B6 complete", lines, ignore.case = TRUE))
    errored    <- any(grepl("Error|Execution halted", lines))
    timelimit  <- any(grepl("DUE TO TIME LIMIT", lines))
    checkpoint <- file.exists(file.path(SCRATCH, "s12v5_ids_checkpoint.rds"))
    cat("Lines logged:", length(lines), "\n")
    cat("Last 10 lines:\n")
    cat(paste(" ", tail(lines, 10), collapse = "\n"), "\n")
    cat("\nStatus:", if (completed)  "COMPLETE — run the load chunk below" else
                    if (timelimit)  paste0("TIME LIMIT — resubmit with sbatch; ",
                                           if (checkpoint) "checkpoint exists, will resume automatically"
                                           else            "no checkpoint found") else
                    if (errored)    "FAILED — check log above for error message" else
                                    "IN PROGRESS", "\n")
} else {
    cat("Log not yet created — job may still be queued.\n")
    cat("Check queue with: squeue --me\n")
}
```

## Load B4–B6 results

Run once the SLURM job is complete, then proceed to B7.

```{r load_b4b6_results}
s12v5.df  <- readRDS(file.path(SCRATCH, "s12v5_df_b6.rds"))
s12v5_ids <- readRDS(file.path(SCRATCH, "s12v5_ids_b6.rds"))
cat(nrow(s12v5.df), "sequences loaded\n")
cat(n_distinct(s12v5.df$taxon), "unique taxa\n")
head(s12v5.df)
```

---

## B4. NCBI GenBank: remote query *(interactive alternative — see SLURM job above)*

> **Note:** The SLURM submission above is recommended for B4–B6. Run the chunks
> below only if running interactively; skip straight to B7 if you used the
> SLURM job.

```{r s12v5_ncbi_query}
cat("Querying NCBI for 12S sequences... this may take several minutes\n")
ncbi.12SV5 <- query_ncbi(marker = "12S", organisms = animals)
cat(length(ncbi.12SV5), "sequences returned from NCBI\n")
```

```{r s12v5_ncbi_pcr}
ncbi.12SV5 <- find_primer_pair(clean(ncbi.12SV5),
                               fwd = s12v5_primers[[1]],
                               rev = s12v5_primers[[2]])
cat(length(ncbi.12SV5), "sequences contain the 12SV5 primer pair\n")

n_before   <- length(ncbi.12SV5)
ncbi.12SV5 <- ncbi.12SV5[!grepl("UNVERIFIED", names(ncbi.12SV5))]
cat("Removed", n_before - length(ncbi.12SV5), "unverified sequences\n")

names(ncbi.12SV5) <-
    names(ncbi.12SV5) |>
    gsub(pattern = " .+$", replacement = "") |>
    gsub(pattern = "^>",   replacement = "")
head(names(ncbi.12SV5))
```

## B5. Combine RefSeq and GenBank

```{r s12v5_combine}
cat("RefSeq sequences: ",  length(refseq.12SV5), "\n")
cat("GenBank sequences:",  length(ncbi.12SV5),   "\n")
cat("Shared accessions: ", intersect(names(ncbi.12SV5), names(refseq.12SV5)) |> length(), "\n")

s12v5.df <- bind_rows(
    data.frame(source = "GenBank", accession = names(ncbi.12SV5),   seq = as.character(ncbi.12SV5)),
    data.frame(source = "RefSeq",  accession = names(refseq.12SV5), seq = as.character(refseq.12SV5))
)
cat(nrow(s12v5.df), "total sequences\n")
```

## B6. Taxonomy lookup

```{r s12v5_taxonomy_lookup}
cat("Looking up taxonomy for", nrow(s12v5.df), "accessions...\n")
s12v5_ids <- accessionToTaxa(s12v5.df$accession, SQL_PATH)
cat("Missing taxonomy IDs:", sum(is.na(s12v5_ids)), "\n")

if (any(is.na(s12v5_ids))) {
    cat("Accessions with missing IDs (assign manually in the next chunk):\n")
    print(s12v5.df[is.na(s12v5_ids), c("source", "accession")])
}
```

```{r s12v5_taxonomy_missing}
# Automatically resolve missing IDs using query_ncbi_accession() for accessions
# added to NCBI after the SQL database was built
if (any(is.na(s12v5_ids))) {
    missing.df <- s12v5.df[is.na(s12v5_ids), c("source", "accession")]
    for (i in seq_len(nrow(missing.df))) {
        idx           <- as.integer(rownames(missing.df)[i])
        acc           <- missing.df[i, 2]
        s12v5_ids[idx] <- query_ncbi_accession(acc)
    }
}

cat("Remaining missing IDs:", sum(is.na(s12v5_ids)), "\n")
```

```{r s12v5_taxonomy_extract}
s12v5_taxonomy <- extract_taxonomy(s12v5_ids, SQL_PATH)
s12v5.df       <- bind_cols(s12v5.df, s12v5_taxonomy)

# Assign lowest available taxonomic rank as display name
s12v5.df <-
    s12v5.df |>
    MButils::lowest_level() |>
    rename(taxon = "name") |>
    select(source, accession, taxon, taxid, superkingdom:forma, seq)

head(s12v5.df)
```

## B6b. Remove off-target sequences

The GenBank query returns any NCBI record matching the 12SV5 primer pair for a queried
species — including off-target hits: viruses co-deposited with host records (e.g. avian
influenza filed under duck species), bacteria from gut-microbiome studies, and synthetic
constructs. In the May 2026 build, 21 such records carried sequences identical to real
fish/bird amplicons and collapsed DADA2's bootstrap to full NA for those taxa.

This deliberately narrows scope relative to the 2025 reference, which carried shrimp,
crab, and jellyfish records: 12SV5 is a vertebrate marker, and those records never
meaningfully assigned (2 ASVs / 15 reads across 385M production reads).

```{r s12v5_offtarget_filter}
# Chordata only — lab decision 2026-08-13. 12SV5 is a vertebrate marker;
# invertebrate reference records proved inert in production data, and edible
# invertebrates are out of assay scope (they were likewise excluded from the
# coverage-recheck candidate lists). Keep this vector in sync with
# ALLOWED_PHYLA in code/qc_reference_build.R.
allowed_phyla_12S <- c("Chordata")

n_before_offtarget <- nrow(s12v5.df)
s12v5.df <- filter(s12v5.df, phylum %in% allowed_phyla_12S)
cat("Off-target filter: removed", n_before_offtarget - nrow(s12v5.df),
    "sequences with phylum outside the vertebrate-marker scope\n")
cat(nrow(s12v5.df), "sequences retained\n")
```

---

## B7. Quality control

```{r s12v5_qc}
s12v5.df$N <- grepl("[AGCT]*[^AGCT]+", s12v5.df$seq)
cat(sum(s12v5.df$N), "sequences with degenerate nucleotides\n")
s12v5.df <- filter(s12v5.df, !s12v5.df$N) |> select(-N)
cat(nrow(s12v5.df), "sequences remaining\n")
```

## B8. Standardise orientation

```{r s12v5_orientation}
fwd_err_12 <- floor(0.2 * length(V5F))
rev_err_12 <- floor(0.2 * length(V5R))
ref_12     <- DNAStringSet(s12v5.df$seq)
names(ref_12) <- paste(s12v5.df$accession, s12v5.df$taxon)

fwd_matches_12 <-
    vmatchPattern(V5F, ref_12, max.mismatch = fwd_err_12, fixed = TRUE) |>
    as.data.frame() |>
    filter(start <= 1) |>
    mutate(type = "forward") |>
    select(group, type)

rev_matches_12 <-
    vmatchPattern(V5R, ref_12, max.mismatch = rev_err_12, fixed = TRUE) |>
    as.data.frame() |>
    filter(start <= 1) |>
    mutate(type = "reverse") |>
    select(group, type)

s12v5.df <-
    bind_rows(fwd_matches_12, rev_matches_12) |>
    arrange(group) |>
    bind_cols(s12v5.df) |>
    mutate(seq = ifelse(type == "reverse",
                        DNAStringSet(seq) |> reverseComplement() |> as.character(),
                        seq)) |>
    select(-c(type, group))

cat("Orientation standardised.\n")
```

## B9. Deduplicate, add controls, and save

```{r s12v5_dedup}
s12v5.df <-
    s12v5.df |>
    group_by(superkingdom, phylum, class, order, family,
             genus, species, subspecies, varietas, forma, taxon, seq) |>
    arrange(desc(source), accession) |>
    summarize(accession = first(accession), .groups = "drop") |>
    arrange(taxon, accession)

cat(nrow(s12v5.df), "sequences after deduplication\n")
cat(n_distinct(s12v5.df$taxon), "unique taxa\n")
```

Append the positive-control spike-ins, exactly as in Part A (see the note at A10). The 12SV5
control is the same physical construct as the trnL one read to a longer amplicon — the 50 bp
trnL sequence is an exact prefix of this 100 bp one.

This step is what was missing previously: the 12SV5 reference shipped without its control, so
spike-in reads were assigned to the nearest real vertebrate (mostly Actinopteri) instead of
being identified as controls.

12SV5 additionally carries a real-species control: *Rhacodactylus leachianus* (New Caledonia
giant gecko, AF090176.1) — a vertebrate the primers amplify but which never occurs in human
diet samples (1.05M reads across 45 samples of the 2025 production megaphyloseq). Every
reference up to 2025 contained it only by inheritance from older broad vertebrate databases;
now that the pipeline builds strictly from `human-foods.csv`, it must be appended here or the
control silently breaks.

```{r s12v5_controls}
controls.12S <-
    read_csv(file.path(REPO_DIR, "data", "inputs", "controls.csv"),
             show_col_types = FALSE) |>
    filter(marker == "12SV5")

if (nrow(controls.12S) > 0) {
    control_rows <-
        controls.12S |>
        # Two kinds of control: synthetic constructs (empty lineage) get the
        # label repeated across superkingdom:species, per the lab convention;
        # real-species controls (e.g. the R. leachianus gecko) carry their true
        # lineage from the lineage column so DADA2 places them correctly.
        mutate(lineage = ifelse(is.na(lineage) | lineage == "",
                                sapply(label, function(l)
                                    paste(c(rep(l, 7), rep("NA", 3)),
                                          collapse = ";")),
                                lineage)) |>
        separate(lineage,
                 into = c("superkingdom", "phylum", "class", "order", "family",
                          "genus", "species", "subspecies", "varietas", "forma"),
                 sep = ";") |>
        mutate(across(superkingdom:forma, \(x) na_if(x, "NA"))) |>
        transmute(superkingdom, phylum, class, order, family, genus, species,
                  subspecies, varietas, forma,
                  taxon = coalesce(species, label), seq = sequence, accession) |>
        # guard against double-adding if this chunk is re-run
        filter(!(seq %in% s12v5.df$seq))

    s12v5.df <- bind_rows(s12v5.df, control_rows)
    cat("Added", nrow(control_rows), "positive-control sequence(s):",
        paste(control_rows$taxon, collapse = ", "), "\n")
} else {
    warning("No 12SV5 controls found in controls.csv - reference will have no positive control.")
}

cat(nrow(s12v5.df), "sequences total (including controls)\n")
```

```{r s12v5_save}
# DADA2: sequence file
s12v5_out <- DNAStringSet(s12v5.df$seq)
names(s12v5_out) <- paste(s12v5.df$accession, s12v5.df$taxon)
writeXStringSet(s12v5_out, file.path(S12V5_DADA2, "12Sv5.fasta"))
cat("Saved 12Sv5.fasta\n")

# DADA2: taxonomy file
s12v5.df_tax     <- s12v5.df |> unite(col = "lineage", superkingdom:forma, sep = ";")
names(s12v5_out) <- s12v5.df_tax$lineage
writeXStringSet(s12v5_out, file.path(S12V5_DADA2, "12Sv5_taxonomy.fasta"))
cat("Saved 12Sv5_taxonomy.fasta\n")

# QIIME2
s12v5.df_tax     <- mutate(s12v5.df_tax, acc = make.unique(accession, sep = "_"))
names(s12v5_out) <- s12v5.df_tax$acc
writeXStringSet(s12v5_out, file.path(S12V5_QIIME2, "12Sv5-sequences.fasta"))
s12v5.df_tax |>
    select(`Feature ID` = acc, Taxon = lineage) |>
    write_delim(file.path(S12V5_QIIME2, "12Sv5-taxonomy.tsv"), delim = "\t")
cat("Saved QIIME2 files\n")

cat("\n=== 12SV5 complete:", nrow(s12v5.df), "sequences,",
    n_distinct(s12v5.df$taxon), "taxa ===\n")
```

---

# Summary

```{r summary}
cat("========================================\n")
cat("FoodSeq Reference Database Build Summary\n")
cat("========================================\n\n")

cat("trnL g/h (plant)\n")
cat("  Sequences:", nrow(trnL.df), "\n")
cat("  Taxa:     ", n_distinct(trnL.df$taxon), "\n")
cat("  Outputs:\n")
cat("   ", file.path(TRNL_DADA2,  "trnLGH.fasta"), "\n")
cat("   ", file.path(TRNL_DADA2,  "trnLGH_taxonomy.fasta"), "\n")
cat("   ", file.path(TRNL_QIIME2, "trnL-sequences.fasta"), "\n")
cat("   ", file.path(TRNL_QIIME2, "trnL-taxonomy.tsv"), "\n\n")

if (exists("trnLCD.df")) {
    cat("trnL c/d (plant)\n")
    cat("  Sequences:", nrow(trnLCD.df), "\n")
    cat("  Taxa:     ", n_distinct(trnLCD.df$taxon), "\n")
    cat("  Outputs:\n")
    cat("   ", file.path(TRNLCD_DADA2,  "trnLCD.fasta"), "\n")
    cat("   ", file.path(TRNLCD_DADA2,  "trnLCD_taxonomy.fasta"), "\n")
    cat("   ", file.path(TRNLCD_QIIME2, "trnLCD-sequences.fasta"), "\n")
    cat("   ", file.path(TRNLCD_QIIME2, "trnLCD-taxonomy.tsv"), "\n\n")
}

cat("12SV5 (vertebrate)\n")
cat("  Sequences:", nrow(s12v5.df), "\n")
cat("  Taxa:     ", n_distinct(s12v5.df$taxon), "\n")
cat("  Outputs:\n")
cat("   ", file.path(S12V5_DADA2,  "12Sv5.fasta"), "\n")
cat("   ", file.path(S12V5_DADA2,  "12Sv5_taxonomy.fasta"), "\n")
cat("   ", file.path(S12V5_QIIME2, "12Sv5-sequences.fasta"), "\n")
cat("   ", file.path(S12V5_QIIME2, "12Sv5-taxonomy.tsv"), "\n")
```
