---
title: "Reference coverage re-check"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

Checks which food species **could be added** to the trnL / 12SV5
reference — without downloading any sequence data itself. Two phases:

-   **Part 1** — of the species already on `human-foods.csv` but missing
    from the current build, which now have marker data at NCBI?
-   **Part 2** (optional, `RUN_PHASE2`) — which globally significant
    foods are missing from `human-foods.csv` entirely?

**Run this before `Extend reference.Rmd`.** This notebook only fires
cheap `esearch` existence checks (no sequence downloads), so it's the
right tool for triage across \~1,500+ species. `Extend reference.Rmd`
does the expensive part — actually pulling and primer-testing each
candidate — so feed it the `CANDIDATES_*` files this notebook writes
rather than pointing it at every still-missing species blind. See
`README.md` ("Extend, or rebuild?").

**A hit here is not a confirmed addition.** A hit count \> 0 only means
NCBI has *some* record mentioning the marker for that species — the
record may not span the primer sites. `Extend reference.Rmd`'s
`find_primer_pair()` is the real test; expect attrition between a
candidate list here and what actually gets added there.

**History.** This consolidates nine standalone scripts that did the same
thing across separate bash + R files, several with paths hardcoded to
one person's machine and one specific past session's temp directory.
Those scripts have been moved to
`archive/code/coverage-recheck/` (`01_...` through `09_...`) for
provenance — they are no longer the way to run this; use this notebook
instead. The probe species lists they used (`probe_staples.txt`,
`probe_extended.txt`) moved with them and are reused as-is by Part 2
below.

# Setup

```{r setup, include = FALSE}
# error = FALSE matters here: knitr::knit() defaults to error = TRUE, which
# records a failed chunk and carries on. A silently-skipped NCBI query would
# leave holes in a checkpointed file that look like "not queried yet" rather
# than "failed" - fail loudly instead.
knitr::opts_chunk$set(echo = TRUE, message = TRUE, error = FALSE)
knitr::opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r libraries, message = FALSE, warning = FALSE}
library(Biostrings)
library(here)
library(rentrez)
library(tidyverse)
```

## Configuration

```{r configuration}
RUN_PHASE1 <- TRUE
RUN_PHASE2 <- TRUE   # FALSE = just re-check known gaps, the routine thing to
                     # do before an Extend reference.Rmd session. Part 2 is
                     # more of an occasional audit against a curated food list.

REF_SUFFIX <- "_Aug2026"   # which built reference to check "missing" against
OUT_SUFFIX <- "_Aug2026"   # suffix for the CSVs this notebook writes

# An NCBI API key lifts the rate limit from 3 to 10 requests/second. Worth
# setting - Part 1 alone fires two rounds of ~1,500 queries.
# rentrez::set_entrez_key("your_key_here")

NCBI_EMAIL <- "your_email@example.com"   # NCBI etiquette: identifies who's querying
NCBI_TOOL  <- "foodseq-coverage-recheck"

OUT_DIR <- here('data', 'outputs', 'coverage-recheck')
RAW_DIR <- file.path(OUT_DIR, 'raw')
dir.create(RAW_DIR, showWarnings = FALSE, recursive = TRUE)

foods <- read_csv(here('data', 'inputs', 'human-foods.csv'), show_col_types = FALSE)
cat(nrow(foods), "species on the target list\n")
```

# NCBI helpers

```{r ncbi-helpers}
# Low-level: hit count for an arbitrary esearch term, retmax=0 so NCBI
# returns just the count. Retried like query_ncbi.R / query_ncbi_accession.R
# elsewhere in this repo, for the same reason - transient NCBI/network
# failures are common across a run this long.
esearch_count <- function(term, max_tries = 3, retry_wait = 2) {
  for (try in seq_len(max_tries)) {
    result <- tryCatch(
      entrez_search(db = "nucleotide", term = term, retmax = 0,
                    email = NCBI_EMAIL, tool = NCBI_TOOL)$count,
      error = function(e) NULL)
    if (!is.null(result)) return(as.integer(result))
    Sys.sleep(retry_wait)
  }
  NA_integer_
}

marker_hit_count <- function(species, marker)
  esearch_count(sprintf('"%s"[Organism] AND %s', species, marker))

# [Organism] is hierarchical, so ANDing Vertebrata[Organism] returns 0 for
# non-vertebrates - resolves vertebrate status in one query rather than a
# separate taxid lookup. 12SV5 is a vertebrate marker; every animal has *a*
# 12S gene, so the unfiltered count alone overstates recoverable candidates.
vertebrate_12S_hits <- function(species)
  esearch_count(sprintf('"%s"[Organism] AND 12S AND Vertebrata[Organism]', species))

# Current NCBI-accepted name for a species. Searching a synonym in the
# Taxonomy database still resolves to the accepted taxon (verified live:
# "Allium porrum" -> "Allium ampeloprasum", taxid 4681) - without this, a
# plain string match against the reference misreports an already-covered
# species as missing just because it's listed under an old name.
resolve_current_name <- function(species, max_tries = 3, retry_wait = 2) {
  for (try in seq_len(max_tries)) {
    result <- tryCatch({
      hit <- entrez_search(db = "taxonomy", term = sprintf('"%s"[All Names]', species),
                            retmax = 1, email = NCBI_EMAIL, tool = NCBI_TOOL)
      if (length(hit$ids) == 0) return(NA_character_)
      entrez_summary(db = "taxonomy", id = hit$ids[1])$scientificname
    }, error = function(e) NULL)
    if (!is.null(result)) return(result)
    Sys.sleep(retry_wait)
  }
  NA_character_
}

# Applies `fun` (species -> single-row tibble) across `species`, checkpointed
# to out_path: a killed or re-run notebook skips whatever's already recorded
# rather than re-querying it. Mirrors the resumability of the original
# 01_ncbi_marker_availability.sh / 02_vertebrate_filter.sh, which mattered
# for a run this size - Part 1 alone is a multi-hour job unauthenticated.
checkpointed_map <- function(species, out_path, fun, rate = 0.4) {
  done <- if (file.exists(out_path)) read_csv(out_path, show_col_types = FALSE) else NULL
  todo <- if (is.null(done)) species else setdiff(species, done$scientific_name)
  if (length(todo) == 0) {
    cat("All", length(species), "species already checkpointed in", basename(out_path), "\n")
    return(read_csv(out_path, show_col_types = FALSE))
  }
  cat("Querying", length(todo), "of", length(species), "species (",
      length(species) - length(todo), "already checkpointed in", basename(out_path), ")...\n")
  pb_step <- max(1, round(length(todo) / 20))
  for (i in seq_along(todo)) {
    row <- fun(todo[i])
    write_csv(row, out_path, append = file.exists(out_path))
    if (i %% pb_step == 0) cat("  ", i, "/", length(todo), "\n")
    Sys.sleep(rate)
  }
  read_csv(out_path, show_col_types = FALSE)
}
```

# Part 1 — re-check species already on the target list

## Current reference coverage

Computed fresh from the built reference each run, not from a saved
"missing species" snapshot - those go stale the moment the reference is
rebuilt, which is exactly what caused 40% of the saved July 2026
candidate list to be outdated names by August (see `README.md`,
"Re-derived against the Aug 2026 build").

```{r phase1-reference, eval = RUN_PHASE1}
ref_paths <- list(
  trnL    = here('data', 'outputs', 'dada2-compatible', 'trnL',
                 paste0('trnLGH_taxonomy', REF_SUFFIX, '.fasta')),
  `12SV5` = here('data', 'outputs', 'dada2-compatible', '12Sv5',
                 paste0('12Sv5_taxonomy', REF_SUFFIX, '.fasta'))
)
stopifnot(file.exists(ref_paths$trnL), file.exists(ref_paths$`12SV5`))

species_in_ref <- function(path) {
  tax <- readDNAStringSet(path)
  sp  <- vapply(strsplit(names(tax), ";"),
                function(x) if (length(x) >= 7) x[7] else NA_character_,
                character(1))
  unique(na.omit(sp))
}

ref_species <- list(trnL = species_in_ref(ref_paths$trnL),
                    `12SV5` = species_in_ref(ref_paths$`12SV5`))
cat("Currently in trnL reference:  ", length(ref_species$trnL), "species\n")
cat("Currently in 12SV5 reference: ", length(ref_species$`12SV5`), "species\n")

plants_missing  <- setdiff(foods$scientific_name[foods$category == "plant"],
                           ref_species$trnL)
animals_missing <- setdiff(foods$scientific_name[foods$category == "animal"],
                           ref_species$`12SV5`)
cat(length(plants_missing),  "food plants missing from the trnL reference\n")
cat(length(animals_missing), "food animals missing from the 12SV5 reference\n")
```

## Resolve current names, and drop anything that's actually already covered

```{r phase1-resolve, eval = RUN_PHASE1}
resolve_and_filter <- function(missing_species, ref_sp, group_label) {
  resolved <- checkpointed_map(
    missing_species,
    file.path(RAW_DIR, paste0('name_resolution_', group_label, OUT_SUFFIX, '.csv')),
    function(sp) tibble(scientific_name = sp, resolved_name = resolve_current_name(sp))
  )
  resolved <- resolved %>%
    mutate(name_changed     = !is.na(resolved_name) & resolved_name != scientific_name,
           already_in_ref   = coalesce(resolved_name, scientific_name) %in% ref_sp,
           query_name       = coalesce(resolved_name, scientific_name))

  n_recovered <- sum(resolved$already_in_ref)
  cat(sprintf("%s: %d/%d resolved names differ from the input; %d were already in the reference under that name (excluded below)\n",
              group_label, sum(resolved$name_changed, na.rm = TRUE), nrow(resolved), n_recovered))

  resolved
}

plants_resolved  <- resolve_and_filter(plants_missing,  ref_species$trnL,    "plants")
animals_resolved <- resolve_and_filter(animals_missing, ref_species$`12SV5`, "animals")

write_csv(filter(plants_resolved, already_in_ref),
          file.path(OUT_DIR, paste0('ALREADY_COVERED_plants_via_synonym', OUT_SUFFIX, '.csv')))
write_csv(filter(animals_resolved, already_in_ref),
          file.path(OUT_DIR, paste0('ALREADY_COVERED_animals_via_synonym', OUT_SUFFIX, '.csv')))

plants_still_missing  <- filter(plants_resolved,  !already_in_ref)
animals_still_missing <- filter(animals_resolved, !already_in_ref)
cat("\nGenuinely still missing - plants:", nrow(plants_still_missing),
    " animals:", nrow(animals_still_missing), "\n")
```

## Query marker availability at NCBI

```{r phase1-query, eval = RUN_PHASE1}
plants_recheck <- checkpointed_map(
  plants_still_missing$query_name,
  file.path(RAW_DIR, paste0('plants_trnL_recheck', OUT_SUFFIX, '.csv')),
  function(sp) tibble(scientific_name = sp, ncbi_hits = marker_hit_count(sp, "trnL"))
)

animals_recheck <- checkpointed_map(
  animals_still_missing$query_name,
  file.path(RAW_DIR, paste0('animals_12S_recheck', OUT_SUFFIX, '.csv')),
  function(sp) tibble(scientific_name = sp, ncbi_hits = marker_hit_count(sp, "12S"))
)

cat("trnL: ", sum(plants_recheck$ncbi_hits > 0, na.rm = TRUE), "/", nrow(plants_recheck),
    " now have some marker data\n", sep = "")
cat("12S:  ", sum(animals_recheck$ncbi_hits > 0, na.rm = TRUE), "/", nrow(animals_recheck),
    " now have some marker data\n", sep = "")
```

## Vertebrate filter (12SV5 only)

12SV5 is a vertebrate marker, but every animal has *a* 12S gene, so the
raw count above overstates recoverable candidates - it will include
shrimp, crab, and other invertebrates the primers cannot amplify.

```{r phase1-vertebrate, eval = RUN_PHASE1}
animals_with_hits <- filter(animals_recheck, ncbi_hits > 0)

vert_check <- checkpointed_map(
  animals_with_hits$scientific_name,
  file.path(RAW_DIR, paste0('animals_vertebrate_check', OUT_SUFFIX, '.csv')),
  function(sp) tibble(scientific_name = sp, vert_12S_hits = vertebrate_12S_hits(sp))
)

animals_vertebrate <- animals_with_hits %>%
  left_join(vert_check, by = "scientific_name") %>%
  mutate(is_vertebrate = vert_12S_hits > 0)

cat(sum(animals_vertebrate$is_vertebrate), "/", nrow(animals_vertebrate),
    " are true vertebrates (12SV5-actionable)\n", sep = "")
```

## Split out higher-rank entries

Some gap-list entries are genus/family/order names, not binomials
(`Siluriformes`, `Mentha`) - `[Organism]` searches expand to all
descendants, so these return large counts that say nothing about
species-level coverage. They need a curation decision (pick a
representative species, or keep as a genus-level entry), not a direct
add.

```{r phase1-rank-split, eval = RUN_PHASE1}
split_by_rank <- function(df) {
  df %>% mutate(n_words = str_count(str_trim(scientific_name), "\\s+") + 1,
                rank_type = if_else(n_words >= 2, "species-level (binomial)",
                                    "higher rank (genus/family/order)"))
}

plants_ranked  <- split_by_rank(plants_recheck)
animals_ranked <- split_by_rank(animals_vertebrate)

higher_rank <- bind_rows(
  transmute(filter(plants_ranked, rank_type != "species-level (binomial)"),
            marker = "trnL", scientific_name, ncbi_hits),
  transmute(filter(animals_ranked, rank_type != "species-level (binomial)"),
            marker = "12S", scientific_name, ncbi_hits = vert_12S_hits)
)
write_csv(higher_rank, file.path(OUT_DIR, paste0('REVIEW_higher-rank-entries', OUT_SUFFIX, '.csv')))
cat(nrow(higher_rank), "higher-rank entries written for manual review\n")
```

## Write candidates

```{r phase1-write, eval = RUN_PHASE1}
plants_candidates <- plants_ranked %>%
  filter(rank_type == "species-level (binomial)", ncbi_hits > 0) %>%
  arrange(desc(ncbi_hits))
write_csv(plants_candidates, file.path(OUT_DIR, paste0('CANDIDATES_plants_trnL', OUT_SUFFIX, '.csv')))
write_csv(filter(plants_ranked, rank_type == "species-level (binomial)", ncbi_hits == 0),
          file.path(OUT_DIR, paste0('STILL_missing_plants_trnL', OUT_SUFFIX, '.csv')))

animals_candidates <- animals_ranked %>%
  filter(rank_type == "species-level (binomial)", is_vertebrate) %>%
  arrange(desc(vert_12S_hits))
write_csv(animals_candidates, file.path(OUT_DIR, paste0('CANDIDATES_animals_12SV5', OUT_SUFFIX, '.csv')))
write_csv(filter(animals_ranked, rank_type == "species-level (binomial)", !is_vertebrate),
          file.path(OUT_DIR, paste0('EXCLUDED_animals_invertebrates', OUT_SUFFIX, '.csv')))
write_csv(filter(animals_recheck, ncbi_hits == 0),
          file.path(OUT_DIR, paste0('STILL_missing_animals_12S', OUT_SUFFIX, '.csv')))

cat("\n=== Part 1 candidates ===\n")
cat("Plants (trnL): ", nrow(plants_candidates), " -> CANDIDATES_plants_trnL", OUT_SUFFIX, ".csv\n", sep = "")
cat("Animals (12SV5): ", nrow(animals_candidates), " -> CANDIDATES_animals_12SV5", OUT_SUFFIX, ".csv\n", sep = "")
cat("\nThese are in the scientific_name-column format code/'Extend reference.Rmd' expects for ADDITIONS.\n")
```

------------------------------------------------------------------------

# Part 2 — foods missing from the target list entirely (optional)

Checks a curated probe list of globally significant foods
(`archive/code/coverage-recheck/probe_staples.txt` + `probe_extended.txt`,
\~199 species weighted toward groups the target list's provenance is
known to be thin on - non-Western livestock, game, camelids) against
`human-foods.csv` membership, then against NCBI for whichever are
absent.

```{r phase2-probe, eval = RUN_PHASE2}
probe <- bind_rows(
  read_delim(here('archive', 'code', 'coverage-recheck', 'probe_staples.txt'), delim = "|",
             col_names = c("scientific_name", "common_name", "group"),
             show_col_types = FALSE, trim_ws = TRUE),
  read_delim(here('archive', 'code', 'coverage-recheck', 'probe_extended.txt'), delim = "|",
             col_names = c("scientific_name", "common_name", "group"),
             show_col_types = FALSE, trim_ws = TRUE)
) %>% distinct(scientific_name, .keep_all = TRUE)
cat(nrow(probe), "species probed\n")

norm <- function(x) x %>% replace_na("") %>% str_to_lower() %>% str_squish()
sci_set   <- norm(foods$scientific_name)
genus_set <- word(sci_set, 1)
hay <- paste(norm(foods$scientific_name), norm(foods$common_name),
             norm(foods$alternative_names), sep = " | ")

probe_status <- probe %>%
  mutate(
    key           = norm(scientific_name),
    genus         = word(key, 1),
    exact_sci     = key %in% sci_set,
    anywhere      = map_lgl(key, ~ any(str_detect(hay, fixed(.x)))),
    genus_present = genus %in% genus_set,
    status = case_when(
      exact_sci     ~ "present (species)",
      anywhere      ~ "present (synonym/alt name)",
      genus_present ~ "GAP (species absent; genus present)",
      TRUE          ~ "GAP (fully absent)"
    ),
    marker = if_else(group == "plant", "trnL", "12S")
  )

cat("\n=== membership status ===\n")
probe_status %>% count(status, sort = TRUE) %>% print(n = Inf)
```

## Check NCBI for the gaps

```{r phase2-ncbi, eval = RUN_PHASE2}
gaps <- filter(probe_status, str_starts(status, "GAP"))
cat(nrow(gaps), "gap species to check\n")

gap_hits <- checkpointed_map(
  gaps$scientific_name,
  file.path(RAW_DIR, paste0('phase2_gap_ncbi', OUT_SUFFIX, '.csv')),
  function(sp) {
    row <- gaps[gaps$scientific_name == sp, ][1, ]
    tibble(scientific_name = sp, ncbi_hits = marker_hit_count(sp, row$marker))
  }
)

phase2_result <- gaps %>%
  left_join(gap_hits, by = "scientific_name") %>%
  mutate(sequence_available = ncbi_hits > 0) %>%
  select(scientific_name, common_name, group, marker, status, ncbi_hits, sequence_available) %>%
  arrange(desc(sequence_available), group, desc(ncbi_hits))

out_path <- file.path(OUT_DIR, paste0('PHASE2_uncatalogued_foods_with_sequence_data', OUT_SUFFIX, '.csv'))
write_csv(phase2_result, out_path)

cat("\n=== Part 2 result ===\n")
cat(sum(phase2_result$sequence_available), "/", nrow(phase2_result),
    " gap species have marker data available -> ", basename(out_path), "\n", sep = "")
cat("\nNOT ready to add to human-foods.csv yet: every existing row carries a\n")
cat("literature/database source. Before acting on this file, add `source` and\n")
cat("`source_status` columns by hand (see PHASE2_uncatalogued_foods_SOURCED.csv\n")
cat("for the format) and save the sourced version separately - do not point\n")
cat("Extend reference.Rmd at the unsourced file directly.\n")
```

------------------------------------------------------------------------

# Summary

```{r summary}
cat("=== Coverage re-check complete ===\n\n")
if (RUN_PHASE1) {
  cat("Part 1 candidates ready for Extend reference.Rmd's ADDITIONS parameter:\n")
  cat("  ", file.path(OUT_DIR, paste0('CANDIDATES_plants_trnL', OUT_SUFFIX, '.csv')), "\n")
  cat("  ", file.path(OUT_DIR, paste0('CANDIDATES_animals_12SV5', OUT_SUFFIX, '.csv')), "\n")
  cat("Review first: REVIEW_higher-rank-entries", OUT_SUFFIX, ".csv (needs a curation decision)\n\n", sep = "")
}
if (RUN_PHASE2) {
  cat("Part 2 needs a manual sourcing pass before anything in it is actionable -\n")
  cat("see PHASE2_uncatalogued_foods_with_sequence_data", OUT_SUFFIX, ".csv\n\n", sep = "")
}
cat("Remember: a hit count here means NCBI has *a* record, not a usable\n")
cat("amplicon. Extend reference.Rmd's find_primer_pair() is what actually\n")
cat("confirms recoverability - expect attrition from these candidate counts.\n")
```
