---
title: "Extend an existing reference database"
output: html_document
---

Adds sequences for a handful of new species to an **existing** reference
without a full cluster rebuild. Runs on a laptop in minutes: it queries NCBI
directly for both sequences and taxonomy, so the ~70 GB `accessionTaxa.sql`
database the full pipeline needs is not required.

**Use this when the only change is adding newly-available species.** Anything
that must alter records already in the reference — a taxonomy fix, a filter
change, a new curation rule — needs a full rebuild, because this notebook only
appends. See `README.md` ("Extend, or rebuild?").

**What it produces.** Both DADA2 files are written: the sequence FASTA
(`>accession taxon`) and the taxonomy FASTA (`>` + 10 semicolon-delimited
ranks). The taxonomy FASTA is the one `assignment_trnL()` / `assignment_12S()`
actually consume, so writing only the sequence file — as the pre-2023 version
of this notebook did — silently fails to extend anything.

**Not for trnLCD.** Part A-CD extracts by alignment against plastome introns,
not by primer match, so extending it needs the panel and belongs in a rebuild.

# Setup

```{r setup, include = FALSE}
# error = FALSE matters here: knitr::knit() defaults to error = TRUE, which
# records a failed chunk and carries on to the next one. This notebook writes
# reference files, so a silently-skipped guard can produce a corrupted output
# that still looks like a clean run. Fail loudly instead.
knitr::opts_chunk$set(echo = TRUE, message = TRUE, error = FALSE)
knitr::opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r libraries, message = FALSE, warning = FALSE}
library(Biostrings)
library(here)
library(rentrez)
library(tidyverse)

source(here('code', 'functions', 'find_primer_pair.R'))
source(here('code', 'functions', 'query_ncbi.R'))
source(here('code', 'functions', 'query_ncbi_accession.R'))
source(here('code', 'functions', 'fetch_lineage_ncbi.R'))
```

## Configuration

Set the marker, point at the reference you are extending, and name the output.
The default output path is deliberately *not* the input path: extending writes
a new versioned pair of files rather than mutating the reference in place.

```{r configuration}
MARKER <- "trnL"          # "trnL" or "12SV5"

# Species to add, one per line, in a CSV with a `scientific_name` column.
# `data/outputs/coverage-recheck/CANDIDATES_*.csv` are in this format.
ADDITIONS <- here('data', 'inputs', 'reference-additions.csv')

IN_SUFFIX  <- "_Aug2026"       # reference being extended
OUT_SUFFIX <- "_Aug2026_ext"   # what to write

# An NCBI API key lifts the rate limit from 3 to 10 requests/second.
# rentrez::set_entrez_key("your_key_here")

stopifnot(MARKER %in% c("trnL", "12SV5"))

cfg <- list(
  trnL = list(
    dir     = here('data', 'outputs', 'dada2-compatible', 'trnL'),
    seq     = "trnLGH%s.fasta",
    tax     = "trnLGH_taxonomy%s.fasta",
    fwd     = DNAString("GGGCAATCCTGAGCCAA"),
    rev     = DNAString("CCATTGAGTCTCTGCACCTATC"),
    query   = "trnL",
    phyla   = c("Streptophyta", "Cyanobacteriota")
  ),
  `12SV5` = list(
    dir     = here('data', 'outputs', 'dada2-compatible', '12Sv5'),
    seq     = "12Sv5%s.fasta",
    tax     = "12Sv5_taxonomy%s.fasta",
    fwd     = DNAString("TAGAACAGGCTCCTCTAG"),
    rev     = DNAString("TTAGATACCCCACTATGC"),
    query   = "12S",
    phyla   = c("Chordata")
  )
)[[MARKER]]

in_seq  <- file.path(cfg$dir, sprintf(cfg$seq, IN_SUFFIX))
in_tax  <- file.path(cfg$dir, sprintf(cfg$tax, IN_SUFFIX))
out_seq <- file.path(cfg$dir, sprintf(cfg$seq, OUT_SUFFIX))
out_tax <- file.path(cfg$dir, sprintf(cfg$tax, OUT_SUFFIX))

stopifnot(file.exists(in_seq), file.exists(in_tax), file.exists(ADDITIONS))
cat("Extending:", basename(in_seq), "+", basename(in_tax), "\n")
cat("Writing:  ", basename(out_seq), "+", basename(out_tax), "\n")
```

# Load the current reference

The two FASTAs are parallel: record *i* of the sequence file is record *i* of
the taxonomy file. That is checked here, because appending to files that have
drifted out of step would corrupt the reference silently.

```{r load-current}
cur_seq <- readDNAStringSet(in_seq)
cur_tax <- readDNAStringSet(in_tax)

cat(length(cur_seq), "records in the sequence FASTA\n")
cat(length(cur_tax), "records in the taxonomy FASTA\n")
stopifnot(length(cur_seq) == length(cur_tax))
# unname() is required: as.character() on a DNAStringSet keeps the names, and
# the two files carry different names by design (accession+taxon vs lineage),
# so comparing them with names attached always reports a spurious mismatch.
stopifnot(identical(unname(as.character(cur_seq)), unname(as.character(cur_tax))))
message("Reference files are parallel and consistent.")

current <- tibble(
  accession = sub("\\s.*$", "", names(cur_seq)),
  taxon     = sub("^\\S+\\s*", "", names(cur_seq)),
  lineage   = names(cur_tax),
  seq       = as.character(cur_seq)
)
```

# Read the species to add

```{r read-additions}
add_list <- read_csv(ADDITIONS, show_col_types = FALSE)
stopifnot("scientific_name" %in% names(add_list))

species <- add_list$scientific_name |> unique() |> na.omit()
cat(length(species), "species requested\n")

already <- species[species %in% current$taxon]
if (length(already) > 0) {
  message(length(already), " already present, skipping: ",
          paste(head(already, 5), collapse = ", "))
  species <- setdiff(species, already)
}
cat(length(species), "species to query\n")
print(species)
```

# Query NCBI and extract the amplicon

Same `query_ncbi()` the full pipeline uses, so the same 50 kb length cap
applies — complete plastid/mitochondrial genomes are excluded. That means a
species whose only record is a whole genome will come back empty here even
though sequence exists; those need a rebuild. Empty results are reported at
the end rather than failing the run.

```{r query-ncbi}
if (length(species) == 0) {
  stop("Nothing to add - every requested species is already in the reference.")
}

raw <- query_ncbi(marker = cfg$query, organisms = species)
cat(length(raw), "records returned\n")

amp <- find_primer_pair(raw, fwd = cfg$fwd, rev = cfg$rev)
cat(length(amp), "contain the", MARKER, "primer pair\n")

amp <- amp[!grepl("UNVERIFIED", names(amp))]
names(amp) <- names(amp) |>
  gsub(pattern = " .+$", replacement = "") |>
  gsub(pattern = "^>",   replacement = "")
cat(length(amp), "after removing UNVERIFIED records\n")
```

# Resolve taxonomy from NCBI

`accession -> taxid -> lineage`, both steps over e-utils. This is the
SQL-free counterpart to the pipeline's `extract_taxonomy()`; it is much slower
per record, which is fine for tens of sequences and would not be for tens of
thousands.

```{r resolve-taxonomy}
accs <- unique(names(amp))
cat("Resolving", length(accs), "accessions to taxon IDs...\n")

taxid_map <- vapply(accs, function(a) {
  id <- query_ncbi_accession(a)
  if (is.null(id) || length(id) == 0) NA_integer_ else as.integer(id)
}, integer(1))

cat(sum(is.na(taxid_map)), "accessions did not resolve\n")

lineages <- fetch_lineage_ncbi(taxid_map[names(amp)])
RANKS <- c("superkingdom","phylum","class","order","family",
           "genus","species","subspecies","varietas","forma")

# Display name = lowest non-NA rank, matching MButils::lowest_level() in the
# pipeline. Reimplemented here so the extension has no MButils dependency.
lowest_level <- function(df) {
  apply(df[, RANKS], 1, function(r) {
    v <- r[!is.na(r) & nzchar(r)]
    if (length(v)) tail(v, 1) else NA_character_
  })
}

new <- tibble(
  accession = names(amp),
  seq       = as.character(amp)
) |>
  bind_cols(lineages[, RANKS]) |>
  mutate(taxon = lowest_level(pick(all_of(RANKS))))

cat(sum(is.na(new$taxon)), "records have no usable taxonomy\n")
```

# QC the additions

The same gates the pipeline applies, in the same order: drop unresolved
taxonomy, degenerate bases, off-target phyla (A6b / B6b), then de-duplicate
against both the additions themselves and the existing reference.

```{r qc-additions}
n0 <- nrow(new)
report <- function(step, df) {
  cat(sprintf("  %-34s %4d -> %4d\n", step, n0, nrow(df)))
  n0 <<- nrow(df)
  df
}
cat("QC:\n")

new <- report("resolved taxonomy + genus", filter(new, !is.na(taxon), !is.na(genus)))
new <- report("no degenerate bases",       filter(new, !grepl("[^ACGT]", seq)))
new <- report("on-target phylum",          filter(new, phylum %in% cfg$phyla))
new <- report("unique within additions",   distinct(new, taxon, seq, .keep_all = TRUE))
new <- report("not already in reference",  anti_join(new, current, by = c("taxon", "seq")))

if (nrow(new) == 0) stop("No additions survived QC - nothing to write.")
```

# Merge and save

```{r merge-save}
merged <- bind_rows(
  current |> select(accession, taxon, lineage, seq),
  new |> transmute(
    accession, taxon,
    lineage = apply(pick(all_of(RANKS)), 1,
                    function(r) paste(ifelse(is.na(r), "NA", r), collapse = ";")),
    seq)
) |>
  arrange(taxon, accession)

cat(nrow(current), "+", nrow(new), "=", nrow(merged), "records\n")

out <- DNAStringSet(merged$seq)

names(out) <- paste(merged$accession, merged$taxon)
writeXStringSet(out, out_seq)

names(out) <- merged$lineage
writeXStringSet(out, out_tax)

cat("Wrote:\n  ", out_seq, "\n  ", out_tax, "\n")
```

# What was added, and what was not

```{r summary}
cat("=== added ===\n")
new |> count(taxon, name = "sequences") |> arrange(desc(sequences)) |> print(n = Inf)

missed <- setdiff(species, new$taxon)
if (length(missed) > 0) {
  cat("\n=== requested but not added ===\n")
  cat("No usable amplicon was recovered for these. Most often the species has\n")
  cat("no sub-50kb record carrying both primer sites - check whether a complete\n")
  cat("genome exists, which this path cannot reach.\n\n")
  print(missed)
}
```

# Verify before shipping

Run the QC gate against the extended pair. It compares them to the reference
you extended, so the record count should rise by exactly what was added and
every other check should pass.

The 4th argument restricts the gate to the marker you extended. Without it the
gate looks for *every* marker under the new suffix and hard-fails on the ones
you did not touch — a pass that reads as a failure.

```{r qc-gate, eval = FALSE}
# From a shell, at the repo root:
#   Rscript code/qc_reference_build.R . _Aug2026_ext _Aug2026 trnL
```

```{r qc-gate-run}
gate_marker <- if (MARKER == "12SV5") "12S" else "trnL"

gate <- system2("Rscript",
                c(shQuote(here("code", "qc_reference_build.R")), shQuote(here()),
                  OUT_SUFFIX, IN_SUFFIX, gate_marker),
                stdout = TRUE, stderr = TRUE)
cat(gate, sep = "\n")

if (!is.null(attr(gate, "status")) && attr(gate, "status") != 0)
  warning("QC gate reported failures - do not ship this reference until they are resolved.")
```
